---
title: "Chapter 8: Tree-based Models"
author: "Katey Yoast"
date: "January 14, 2016"
output:
  html_document:
    keep_md: yes
---

##Introduction
Tree-based models are a supervised machine learning method commonly used in soil survey and ecology for exploratory data analysis and prediction due to their simplistic nonparametric design. Instead of fitting a model to the data, tree-based models recursively partition the data into increasingly homogenous groups based on values that minimize a loss function (such as Sum of Squared Errors (SSE) for regression or Gini Index for classification) (McBratney et al., 2003). The two most common packages for generating tree-based models in R are rpart and randomForest. The rpart package creates a regression or classification tree based on binary splits that maximize homogeneity and minimize impurity. The output is a single decision tree that can be further "pruned" or trimmed back using the cross-validation error statistic to reduce over-fitting. The randomForest package is similar to rpart, but is double random in that each node is split using a random subset of predictors AND observations at each node and this process is repeated hundreds of times (as specified by the user). Unlike rpart, random forests do not produce a graphical decision tree since the predictions are averaged across hundreds or thousands of trees. Instead, random forests produce a variable importance plot and a tabular statistical summary. 
<br/>

##8.1 Classification and Regression Trees (CART)
The basic function for all CART models is (y ~ x), where y is the dependent variable to be predicted from x, a set of independent variables. If the dependent variable (y) is numeric, the resulting tree will be a regression tree. Conversely, if the dependent variable (y) is categorical, the resulting tree will be a classification tree. The rpart package allows all data types to be used as independent variables, regardless of whether the model is a classification or regression tree. The rpart algorithm ignores missing values when determining the quality of a split and uses surrogate splits to determine if observation(s) with missing data is best split left or right. If an observation is missing all surrogate splits, then the observation(s) is sent to the child node with the largest relative frequency (Feelders, 1999).

Assuming that the rpart and randomForest packages are already installed on your machine, simply load the packages using the library( ) function.

```{r, warning=FALSE, message=FALSE}
library(rpart)
library(randomForest)

file<-'https://raw.githubusercontent.com/ncss-tech/stats_for_soil_survey/master/data/logistic/wv_transect_editedforR.csv'
download.file(file, destfile = "soildata.csv")
soildata<-read.csv("soildata.csv", header=TRUE, sep=",")


```

Review the data structure to ensure that the data were imported correctly into R.
```{r}
str(soildata)

```

The example dataset, soildata, consists of 250 observations and 58 variables that were collected in the field or derived from geospatial data to identify spodic soil properties in West Virginia. Of particular interest is determining best splits for spodic intensity (relative spodicity index). As you can see in that data structure, R interpreted the spodint field as numeric. Since spodint is an index, it will need to be changed to a factor and then to an ordered factor. 

```{r}
soildata$spodint<-as.factor(soildata$spodint)
soildata$spodint<-ordered(soildata$spodint)

```

If you wanted to create a classification tree for spodint using all of the variables, you would simply type: rpart(spodint ~ ., data=soildata). Since the soildata dataset contains variables such as FID and NASIS_Pedo that are site observation labels, it is not wise to include all variables in the model. Instead, you want to define which variables the model should use. 

```{r, fig.width=7, fig.height=8}
spodintmodel<-rpart(spodint~x+y+Overtype+Underconifer+Oi+Oe+Oa+Otot+epipedon+subgroup+order+ps+drainage+slope+surfacetex+stoniness+depthclass+bedrockdepth+hillslope+tipmound+rainfall+geology+aachn+dem10m+downslpgra+eastness+greenrefl+landsatb1+landsatb2+landsatb3+landsatb7+maxc100+maxent+minc100+mirref+ndvi+northeastn+northness+northwestn+planc100+proc100+protection+relpos11+slp50+solar+tanc75, data=soildata, method = "class")

spodintmodel

plot(spodintmodel)
text(spodintmodel, cex=0.8) ##cex is text size

```

When soil order and subgroup are included in the spodint model, they are the two most important variables for seperating spodic intensity. This makes sense given the ratings for spodic intensity: 0 = non spodic, 0.5 = Bs horizon, 1 = spodic subgroup, 1.5 = spodic subgroup OR Spodosol, 2 = Spodosol. Let's omit soil order and subgroup to see what other factors best seperate spodint.  

```{r, fig.width=7, fig.height=8}
spodintmodel2<-rpart(spodint~x+y+Overtype+Underconifer+Oi+Oe+Oa+Otot+epipedon+ps+drainage+slope+surfacetex+stoniness+depthclass+bedrockdepth+hillslope+tipmound+rainfall+geology+aachn+dem10m+downslpgra+eastness+greenrefl+landsatb1+landsatb2+landsatb3+landsatb7+maxc100+maxent+minc100+mirref+ndvi+northeastn+northness+northwestn+planc100+proc100+protection+relpos11+slp50+solar+tanc75, data=soildata, method = "class")

spodintmodel2

plot(spodintmodel2)
text(spodintmodel2, cex=0.8)

```

The decision tree got more complex when soil order and subgroup were omitted. Also, 0.5 and 1.5 were omitted from the terminal nodes. One way to compare the two models is to use the function: printcp ( ).

```{r}
printcp(spodintmodel)
printcp(spodintmodel2)

```

The printcp ( ) funtion generates a cost complexity parameter table that provides the complexity parameter value (CP), relative model error (1 - relative error = ~variance explained), error estimated from a 10-fold cross validation (xerror), and the standard error of the xerror (xstd). The CP values control the size of the tree; the greater the CP value, the fewer the number of splits in the tree. To determine the optimal CP value, rpart automatically performs a 10-fold cross validation. The optimal size of the tree is generally the row in the CP table that minimizes all error with the fewest branches. Another way to determine the optimal tree size is to use the plotcp ( ) function. This will plot the xerror versus cp value and tree size. 

```{r}
plotcp(spodintmodel)
plotcp(spodintmodel2)

```

The optimal CP value is 0.099 for spodintmodel and 0.034 for spodintmodel2. Since both spodic intensity models overfit that data, they will need to be pruned using the prune ( ) function.

```{r, fig.width=7, fig.height=8}
pruned<-prune(spodintmodel, cp=0.099)
printcp(pruned)
plot(pruned)
text(pruned, cex=0.8)

pruned2<-prune(spodintmodel2, cp=0.034)
printcp(pruned2)
plot(pruned2)
text(pruned2, cex=0.8)

```

Spodintmodel explained approximately 83% of the variance in the data while spodintmodel2 only explained 41%. It is evident that soil order and subgroup greatly influence spodic intensity.  

<br/>

##8.2 Random Forest
The randomForest algorithm fits hundreds to thousands of CART models to random subsets of input data and combines the trees for prediction. Similarly to rpart, randomForest allows all data types to be used as independent variables, regardless of whether the model is a classification or regression tree. Unlike rpart, the randomForest algorithm does not straight forwardly handle missing values with surrogate splits. There is a function called rfImpute ( ) that uses a proximity matrix from the randomForest to populate missing values with either the weighted average of the non-missing observations (weighted by the proximities) for continuous predictors or the category with the largest average proximity for categorical predictors. 

Going back to the soildata dataset, it is also of particular interest to determine what properties best predict total O horizon thickness (Otot). Just like rpart, randomForest has the same basic model function: (y ~ x).

```{r}
rf<-randomForest(Otot~x+y+Overtype+Underconifer+spodint+ps+drainage+slope+surfacetex+stoniness+depthclass+bedrockdepth+hillslope+tipmound+rainfall+geology+aachn+dem10m+downslpgra+eastness+greenrefl+landsatb1+landsatb2+landsatb3+landsatb7+maxc100+maxent+minc100+mirref+ndvi+northeastn+northness+northwestn+planc100+proc100+protection+relpos11+slp50+solar+tanc75, data=soildata, importance=TRUE) 
##Oi, Oe, Oa, order, subgroup, and epipedon were omitted as independent variables
##importance=TRUE will allow the generation of a variable importance plot

rf ## statistical summary

plot(rf)  ##out of bag (OOB) error rate versus number of trees


```

The rf model, generated using the default number of trees and number of variables tried at each split, explained approximately 40% of the variance and produced a mean square error (sum of squared residuals divided by n) of 21 cm2. Another way to assess the rf model is to look at the variable importance plot. 

```{r, fig.width=7, fig.height=8}
varImpPlot(rf)

importance(rf) ##tabular summary

```

For each tree, each predictor in the OOB sample is randomly permuted and passed through the tree to obtain an error rate (mean square error (MSE) for regression and Gini index for classification). The error rate from the unpermuted OOB is then subtracted from the error rate of the permuted OOB data, and averaged across all trees. When this value is large, it implies that a variable is highly correlated to the dependent variable and is needed in the model. 

In a regression tree analysis, randomForest uses %IncMSE and IncNodePurity to rank variable importance. %IncMSE is simply the average increase in squared residuals of the test set when variables are randomly permuted (little importance = little change in model when variable is removed or added) and IncNodePurity is the increase in homogeneity in the data partitions. In a classification tree analysis, randomForest uses MeanDecreaseAccuracy and MeanDecreaseGini. For MeanDecreaseAccuracy, the more the accuracy of the model decreases due to the addition of a single variable, the more important the variable is deemed. MeanDecreaseGini is a measure of how each variable contributes to the homogeneity of the nodes and leaves.

In the rf model, it is apparent that spodint is the most important variable used in the model, followed by y, landsatb7, Overtype, surfacetex, maxent, hillslope, slp50, rainfall, and protection index. 

<br/>

##8.3 Prediction using Tree-based Models
As with any modeling technique, tree-based models can be used for prediction and can be spatially interpolated using environmental covariates. Most commonly, whole datasets are subdivided into train and test data sets:

```{r}
## splits 70% of the data selected randomly into training set and the remaining 30% sample into test data set
datasplit<-sort(sample(nrow(soildata), nrow(soildata)*.7)) 
train<-soildata[datasplit,]
test<-soildata[-datasplit,]

str(train)
str(test)

```

