
<!-- You will need to the following packages: -->
<!--    * `leafsync`: synchronized mapview panels -->
<!--    * `rasterVis`: fancy plotting for `raster` objects -->

<!-- Use the following as a template to tinker with some gridded soil property data (800m grid, derived from SSURGO/STATSGO). See `?ISSR800.wcs` for details. -->

<!-- The [soilDB WCS tutorial](http://ncss-tech.github.io/AQP/soilDB/WCS-demonstration-01.html) contains many more examples from which to draw ideas. -->

<!-- ```{r fig.width} -->
<!-- # WCS interface -->
<!-- library(soilDB) -->
<!-- # wrangling polygons -->
<!-- library(sp) -->
<!-- library(sf) -->
<!-- # raster data visualization -->
<!-- library(rasterVis) -->

<!-- # make a bounding box and assign a CRS (4326: GCS, WGS84) -->
<!-- bb <- st_bbox( -->
<!--   c(xmin = -121, xmax = -120, ymin = 37, ymax = 38),  -->
<!--   crs = st_crs(4326) -->
<!-- ) -->

<!-- # convert bbox to sf geometry -->
<!-- bb <- st_as_sfc(bb) -->

<!-- # try some others from the complete list of available grids: -->
<!-- # WCS_details(wcs = 'ISSR800') -->

<!-- # get soil texture class, 0-25cm -->
<!-- r1 <- ISSR800.wcs(aoi = bb, var = 'texture_025cm') -->

<!-- # get pH, 0-25cm -->
<!-- r2 <- ISSR800.wcs(aoi = bb, var = 'ph_025cm') -->


<!-- # grid of category IDs + labels stored in a  -->
<!-- # raster attribute table (RAT) -->
<!-- r1 -->

<!-- # grid of continuous values -->
<!-- r2 -->

<!-- # plot -->
<!-- levelplot(r1, margin = FALSE) -->
<!-- levelplot(r2, margin = FALSE) -->
<!-- ``` -->


<!-- Sync-ed `mapview` panels. This can be an effective way to explore similarities or differences between two spatial datasets. Try adjusting the `alpha.regions` argument (value range: 0-1) to `mapview()` to set transparency (lower values = more transparent). -->
<!-- ```{r out.width="100%"} -->
<!-- # interactive mapping -->
<!-- library(mapview) -->
<!-- # sync-ed panels -->
<!-- library(leafsync) -->

<!-- # try  -->
<!-- m1 <- mapview(r1, map.types = "Esri.WorldImagery", legend = TRUE, na.color = NA) -->
<!-- m2 <- mapview(r2, map.types = "Esri.WorldImagery", legend = TRUE, na.color = NA) -->

<!-- # build synced panels. -->
<!-- sync(m1, m2) -->
<!-- ``` -->


<!-- Make another pair of synchronized `mapview` panels, this time using SSURGO and STATSGO map unit delineations via SDA. This example uses the [`SDA_spatialQuery()`](http://ncss-tech.github.io/soilDB/reference/SDA_spatialQuery.html) function to request vector data from SDA using a template `sp` object. The `geomIntersection = TRUE` argument to `SDA_spatialQuery()` causes SDA to compute the spatial intersection between template (bounding box) and map unit delineations. SDA has fairly tight constraints on how much data can be returned per request (32Mb), so you will have to plan carefully when working with large areas of interest. -->

<!-- ```{r out.width="100%"} -->
<!-- # make a smaller bounding box -->
<!-- bb <- st_bbox(c(xmin = -114.16, xmax = -114.08, ymin = 47.65, ymax = 47.68), crs = st_crs(4326)) -->
<!-- # convert to sf geometry -->
<!-- bb <- st_as_sfc(bb) -->

<!-- ## TODO: SDA_spatialQuery() currently expects an sp object -->
<!-- # convert to sp object -->
<!-- bb <- as_Spatial(bb) -->

<!-- # get SSURGO and STATSGO -->
<!-- # result is a SpatialPolygonsDataFrame -->
<!-- ssurgo <- SDA_spatialQuery(bb, what = 'geom', geomIntersection = TRUE) -->
<!-- statsgo <- SDA_spatialQuery(bb, what = 'geom', geomIntersection = TRUE, db = 'STATSGO') -->

<!-- # make mapview panels -->
<!-- # adjust line styles, disable legends -->
<!-- m1 <- mapview(ssurgo, map.types = "Esri.WorldImagery", legend = FALSE, na.color = NA, color = 'yellow', fill = NA, lwd = 1) -->
<!-- m2 <- mapview(statsgo, map.types = "Esri.WorldImagery", legend = FALSE, na.color = NA, color = 'white', fill = NA, lwd = 1) -->

<!-- # sync -->
<!-- sync(m1, m2) -->
<!-- ``` -->

<!--
While it is possible to hand-make `sp` class objects, it is usually faster to import existing raster or vector files from disk. More on this later. For the sake of demonstration we will make a `SpatialPointsDataFrame` object. This is analogous to the point type feature class or shapefile.

```{r}
# create point geometry from coordinates
p <- SpatialPoints(coords = cbind(-97.721210, 40.446068))
# attach attribute table
p <- SpatialPointsDataFrame(p, data=data.frame(id=1, taxonname='alpha'))
# check internal structure
str(p)
```

This is an "S4" object with "slots" (e.g. `@data`) that are used to store various components. Most all S4 objects have specialized functions for getting and setting the contents of the slots. For example, the `bbox()` function is used to get or set the contents of the `@bbox` slot.

There is a convenient shortcut for "upgrading" point data that is stored in a `data.frame`. This is especially useful for preparing NASIS/KSSL data for spatial operations.

```{r}
# make some fake data
d <- data.frame(x=runif(10), y=runif(10), id=1:10, code=letters[1:10])
# upgrade to SpatialPointsDataFrame
coordinates(d) <- ~ x + y
# check
class(d)
```

See the [sp gallery](https://edzer.github.io/sp/) and [this vignette](https://cran.r-project.org/web/packages/sp/vignettes/intro_sp.pdf) for a much more detailed description of the `sp` classes.

#### Interacting with `sp` Objects

`sp` objects are modeled after `data.frame` objects; rows (features) and columns (attributes) are accessed using `[]` or `$` notation.

Extract the first 5 features (and all associated attributes) from a `SpatialPointsDataFrame` object:
```{r }
# [rows, columns]
# [features, attributes]
d[1:5, ]
```

Extract the attributed called `id`:
```{r}
d$id
```

Create some random numbers and save to a new attribute called `rand`:
```{r}
# new data must be of the same length as number of features, otherwise recycling will occur
d$rand <- rnorm(n=nrow(d))
```

Filter features based on attributes:
```{r}
idx <- which(d$rand > 0)
d[idx, ]
```

Convert back to a `data.frame` object:
```{r}
as(d, 'data.frame')
```

#### Coordinate Reference Systems and `proj4` Syntax

Spatial data aren't all that useful without an accurate description of the coordinate reference system (CRS). This type of information is typically stored within the ".prj" component of a shapefile, or in the header of a GeoTiff. Without a CRS it is not possible to perform coordinate transformations (e.g. conversion of geographic coordinates to projected coordinates), spatial overlay (e.g. intersection), or geometric calculations (e.g. distance or area).

In R, CRS information is encoded using ["proj4" notation](http://proj4.org/), which looks something like this: `'+proj=longlat +datum=WGS84'` (geographic coordinates referenced to the WGS84 datum). You can look-up CRS definitions in many different formats on the [spatialreference.org](http://spatialreference.org/) website. Some examples:
  
  * `+proj=longlat +datum=NAD83`: geographic coordinate system, NAD83 datum
* `+proj=longlat +datum=NAD27`: geographic coordinate system, NAD27 datum
* `+proj=utm +zone=10 +datum=NAD83`: projected coordinate system (UTM zone 10), NAD83 datum
* `+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23.0 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs`: AEA CONUS / gSSURGO

Returning to the example from above, lets **assign** a CRS to this point using the `proj4string()` function. Note that this process doesn't alter the geometry of the object.
```{r }
proj4string(p) <- '+proj=longlat +datum=WGS84' 
str(p)
```

Transformation of points, lines, and polygons is a simple matter--as long the source and destination CRS data are available.
```{r}
# transform to UTM zone 14
p.utm <- spTransform(p, CRS('+proj=utm +zone=14 +datum=NAD83'))
cbind(gcs=coordinates(p), utm=coordinates(p.utm))

# transform to GCS NAD27
p.nad27 <- spTransform(p, CRS('+proj=longlat +datum=NAD27'))
cbind(coordinates(p), coordinates(p.nad27))
```

CRS definitions can also be generated from [EPSG](http://epsg.io/) codes. For example, a geographic coordinate system referenced to the WGS84 datum can be accessed using EPSG code 4326:
```{r}
CRS('+init=epsg:4326')
```

### Accessing Spatial Data Sources via GDAL/OGR

Many of the vector data sources we use can be directly translated to `sp` class objects via `rgdal` package. The `rgdal` package provides linkages to the powerful [GDAL/OGR](http://www.gdal.org/) library, the details of which are well beyond the scope of this training. Suffice to say, `rgdal` makes it possible to import just about any vector or raster data using two basic functions:

 * `readOGR()`: vector data sources such as shape files, KML, etc.
 * `readGDAL()`: raster data sources such as GeoTiff, IMG, ArcGRID, ASCII grids, etc.
 
A full listing of available drivers can be accessed with `ogrDrivers()` and `gdalDrivers()`. Note that it is not yet possible to load raster data from an ESRI File Geodatabase.

This module will not cover importing or exporting raster data via `rgdal` as the `raster` package provides a much more convenient interface to the GDAL library.
-->

## EXERCISE: Working with real data



Quick inspection of the data.

```{r, eval=TRUE}
# object class
class(mlra)
class(maat)
class(rs)

# the raster package provides a nice "print" method for raster and sp classes
print(maat)

# coordinate reference systems: note that they are not all the same
proj4string(mlra)
proj4string(maat)
proj4string(rs)
```

Basic plot methods (class-specific functions) for the data. Note that this approach requires that all layers in the "map" are in the same coordinate refrence system (CRS).


Try the `rasterVis` package approach to plotting `raster` and `sp` objects. Syntax is more expressive, figures are more intricate, but there is a learning curve. See the [rasterVis tutorial](https://oscarperpinan.github.io/rastervis/) for a comprehensive set of examples.
```{r fig.width=5, fig.height=6.75}
levelplot(
  # raster object
  maat,
  # suppress marginal plots
  margin = FALSE,
  # custom colors
  par.settings = BuRdTheme,
  # combine multiple spatial data layer here
  panel = function(...) {
    panel.levelplot(...)
    sp.lines(mlra.gcs, col = 'black')
  }
  )
```