---
title: Chapter 4 - Exploratory Data Analysis
author: Stephen Roecker and Tom D'Avello
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    css: css/hpstr.css
    fig_caption: yes
    fig_height: 3
    fig_width: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(warning=FALSE, message=FALSE)

library(aqp)
library(soilDB)
library(ggplot2)

```


## Objectives

* Review methods for estimating Low, RV, and High values
* Review different methods for visualing soil data


---


## Range in Characteristics (RIC)

* Why does the L, RV, and H clay represent?
    + What taxonomic unit do they refer too?
* What is the L and H for texture?

```{r, echo=FALSE}

crosby <- fetchSDA(WHERE = "muname LIKE 'Crosby%' AND majcompflag = 'Yes'")
spc <- crosby
c2 <- subsetProfiles(spc, s = paste0("cokey == ", sample(spc, 1)@site$cokey))
c_h <- horizons(c2)
names(c_h) <- gsub("total", "", names(c_h)) 

vars <- c("hzname", "clay_l", "clay_r", "clay_h", "texture")
knitr::kable(c_h[vars], row.names = FALSE)

```


---


##  Descriptive Statistics - Estimates of L, RV & H

```{r, echo=FALSE}

desc <- data.frame(
  Parameter = c("Mean", "Median", "Mode", "Standard Deviation", "Quantiles"),
  NASIS = c(rep("RV", 3), rep("L & H", 2)),
  Description = c("arithmetic average", 
                  "middle value, 50% quantile", 
                  "most frequent value", "variation", 
                  "percent rank of values, such that all values are <= p"
                  ),
  'R function' = c("mean()", "median()", "sort(table(), decreasting = TRUE)[1]", "sd()", "quantile()"),
  check.names = FALSE
  )
knitr::kable(desc)

```
 

---


## Descriptive Statistics - Numeric Data

```{r}
data("loafercreek")
h <- horizons(loafercreek) 
h$texture_class <- factor(h$texture_class)

vars <- c("clay", "phfield", "total_frags_pct", "texture_class")
summary(h[vars])

```


---


## Descriptive Statistics - Categorical Data

```{r}

table(h$genhz, h$texture_class, useNA = "ifany")

```


---


## Data Inspection - Missing Data

 1. **Exclude** all rows or columns that contain missing values using the function `na.exclude()`, such as `h2 <- na.exclude(h)`. However this can be wasteful because it removes all rows (e.g., horizons), regardless if the row only has 1 missing value. Instead it's sometimes best to create a temporary copy of the variable in question and then remove the missing variables, such as `clay <- na.exclude(h$clay)`.
 2.	**Replace** missing values with another value, such as zero, a global constant, or the mean or median value for that column, such as `h$clay <- ifelse(is.na(h$clay), 0, h$clay) # or h[is.na(h$clay), ] <- 0`.
 3. **Read** the help file for the function you're attempting to use. Many functions have additional arguments for dealing with missing values, such as `na.rm`.


---


## Graphical Methods - Descriptions

```{r, echo=FALSE}

figs <- data.frame(
  'Plot Types' = c("Bar", "Histogram", "Density", "Quantile-Quantile", "Box-Whisker", "Scatter & Line"),
   Description = c("a plot where each bar represents the frequency of observations for a 'group'",
                   "a plot where each bar represents the frequency of observations for a 'given range of values'",
                  "an estimation of the frequency distribution based on the sample data",
                  "a plot of the actual data values against a normal distribution",
                  "a visual representation of median, quartiles, symmetry, skewness, and outliers",
                   "a graphical display of one variable plotted on the x axis and another on the y axis"
                  ),
  check.names = FALSE
  )

knitr::kable(figs)

```


---
 

## Graphical Methods - Functions

```{r, echo=FALSE}

figs <- data.frame(
  'Plot Types' = c("Bar", "Histogram", "Density", "Quantile-Quantile", "Box-Whisker", "Scatter & Line"),
  # Description = c("a bar plot where each bar represents the frequency of observations for a given range of values",
  #                "an estimation of the frequency distribution based on the sample data",
  #                "a plot of the actual data values against a normal distribution",
  #                "a visual representation of median, quartiles, symmetry, skewness, and outliers",
  #                 "a graphical display of one variable plotted on the x axis and another on the y axis",
  #                "plots formatted for the representation of circular data"
  #                ),
  'Base R' = c("barplot()", "hist()", "plot(density())", "qqnorm()", "boxplot()", "plot()"),
  'lattice' = c("barchart()", "histogram()", "densityplot()", "qq()", "bwplot()", "xyplot"),
  'ggplot geoms' = c("geom_bar()", "geom_histogram()", "geom_density()", "geom_qq()", "geom_boxplot()", "geom_point()"),
  check.names = FALSE
  )

knitr::kable(figs)

```


---


## Distributions

```{r, echo=FALSE, fig.dim = c(8, 4)}

data("metadata")

# Generalized the horizon designations
n <- c("A",
       "BAt",
       "Bt1",
       "Bt2",
       "Cr",
       "R")
# REGEX rules
p <- c("A",
       "BA|AB",
       "Bt|Bw",
       "Bt3|Bt4|2B|C",
       "Cr",
       "R")

loafercreek$genhz <- generalize.hz(loafercreek$hzname, n, p)

h <- horizons(loafercreek)
h <- within(h, {
  hzdepm = (hzdept + hzdepb) / 2
  texture_class = factor(texture_class, 
                         levels = metadata[metadata$ColumnPhysicalName == "texcl", "ChoiceName"]
                         )
  })

idx <- is.na(h$texture_class)

# bar plot
p_b <- ggplot(h[!idx, ], aes(x = texture_class)) +
  geom_bar() +
  ylab("count (n)") + xlab("texture class") +
  theme(aspect.ratio = 1) +
  ggtitle("Bar Plot")

# histogram
p <- ggplot(data = h, aes(x = clay))

p_h <- p + 
  geom_histogram() + 
  xlab("clay (%)") + ylab("count (n)") +
  theme(aspect.ratio = 1) +
  ggtitle("Histogram")

# density plot
p_d <- p + 
  geom_density(fill = "grey", alpha = 0.5) + 
  xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Density Plot")

gridExtra::grid.arrange(p_b, p_h, p_d, ncol = 3)

```


---


## The 'Normal Distribution

```{r, echo = FALSE, fig.dim = c(8, 4)}

dnorm_limit_1 <- function(x) {
  y <- dnorm(x)
  y[-1 > x | x > 1] <- NA
  return(y)
  }
dnorm_limit_2 <- function(x) {
  y <- dnorm(x)
  y[-2 > x | x > 2] <- NA
  return(y)
  }

# ggplot() with dummy data
ggplot(data.frame(x=c(-3, 3)), aes(x=x)) + 
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm_limit_1, geom="area", fill="blue", alpha=0.2) +
  stat_function(fun=dnorm_limit_2, geom="area", fill="orange", alpha=0.2) +
  annotate("text", x = 0, y = 0.2, label = "1 sd = 68%") +
  annotate("text", x = -1.5, y = 0.03, label = "2 sd = 95%")

```


---


## Mean vs Median


```{r, echo=FALSE, fig.dim = c(8, 4)}

p   <- c(0.025, 0.25, 0.5, 0.75, 0.975)

avg <- mean(h$clay, na.rm = TRUE)
std <- sd(h$clay,   na.rm = TRUE)

clay <- rbind(
  data.frame(
    value = c(avg - 2 * std, avg, avg + 2 * std),
    variable = "mean & sd",
    stringsAsFactors = TRUE
    ),
  data.frame(
    value    = quantile(h$clay, p, na.rm = TRUE),
    variable = "quantiles",
    stringsAsFactors = TRUE
    )
  )

avg <- mean(h$total_frags_pct, na.rm = TRUE)
std <- sd(h$total_frags_pct,   na.rm = TRUE)

frags <- rbind(
  data.frame(
    variable = "mean & sd",
    value = c(avg - 2 * std, avg, avg + 2 * std),
    stringsAsFactors = TRUE
    ),
  data.frame(
    variable = "quantiles", 
    value    = quantile(h$total_frags_pct, p, na.rm = TRUE),
    stringsAsFactors = TRUE
    )
  )

p1 <- ggplot(h, aes(x = clay)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(data = clay, aes(xintercept = value, lty = variable)) +
  xlab("%") +
  #xlim(-5, max(h$total_frags_pct)) +
  theme(aspect.ratio = 1) +
  ggtitle("Clay")

p2 <- ggplot(h, aes(x = total_frags_pct)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(data = frags, aes(xintercept = value, lty = variable)) +
  xlab("%") +
  #xlim(-5, max(h$total_frags_pct)) +
  theme(aspect.ratio = 1) +
  ggtitle("Rock Fragments")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```


---


## Distributions

```{r, echo=FALSE, fig.dim = c(8, 4)}

# qq plot
p_qq <- ggplot(h, aes(sample = clay)) + 
  geom_qq() +
  geom_qq_line() +
  xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Q-Q Plot")

# box plot
p_box <- ggplot(h, aes(x = 1, y = clay)) + 
  geom_boxplot() +
  ylab("clay (%)") +
  theme(aspect.ratio = 1) +
  coord_flip() +
  ggtitle("Box Plot")

gridExtra::grid.arrange(p_box, p_qq, ncol = 2)

```


---


## Relationships

```{r, echo=FALSE, fig.dim = c(8, 4)}

# scatter plot
p_s <- ggplot(h, aes(x = clay, y = hzdepm)) +
  geom_point() +
  ylim(100, 0) +
  ylab("depth (cm)") + xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Scatter Plot (Soil Horizons)")

l_p <- ggplot(h) +
  # geom_smooth(aes(y = clay, x = hzdepm), se = FALSE) +
  geom_line(aes(y = clay, x = hzdepm, group = peiid)) +
  xlim(100, 0) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1) +
  ggtitle("Line Plot (Soil Profiles)")

# l_s <- ggplot(h, aes(y = clay, x = hzdepm, group = peiid)) +
#   geom_step(aes(y = clay, x = hzdept, group = peiid), direction = "vh") +
#   xlim(100, 0) +
#   xlab("depth (cm)") + ylab("clay (%)") +
#   coord_flip() +
#   theme(aspect.ratio = 1) +
#   ggtitle("Step Line Plot (Soil Profiles)")
# 
# test3 <- aqp::slice(loafercreek, seq(0, 100, 2) ~ clay)
# test3 <- horizons(test3)
# test3$hzdepm <- with(test3, (hzdepb + hzdept) / 2)
# 
# l_sm <- ggplot(test3, aes(y = clay, x = hzdept, group = peiid)) +
#   geom_line(stat = "smooth", method = "loess", col = "black", alpha = 0.5) +
#   xlim(100, 0) +
#   xlab("depth (cm)") + ylab("clay (%)") +
#   coord_flip() +
#   theme(aspect.ratio = 1) +
#   ggtitle("Line Plot (Soil Profiles)")


gridExtra::grid.arrange(p_s, l_p, ncol = 2)

```


---


## 3rd Dimension - Color, Shape, Size, Layers, etc...

```{r, echo=FALSE, fig.dim = c(8, 4)}
h$clay2 <- ifelse(is.na(h$clay), 0, h$clay)

ggplot(h[!idx, ], aes(y = clay, x = hzdepm, col = genhz)) +
  geom_point(size = 2) +
  geom_smooth(se = FALSE, col = "black") +
  xlim(100, 0) + ylim(min(h$clay, na.rm = TRUE), max(h$clay, na.rm = TRUE)) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1)

```


---


## 3rd Dimension - Color

```{r, echo=FALSE, fig.dim = c(8, 4)}

h$genhz2 <- factor(h$genhz, levels = rev(levels(h$genhz)))

p_sb1 <- ggplot(h, aes(x = genhz2, fill = texture_class)) +
  geom_bar(position = "stack") +
  theme(aspect.ratio = 1) +
  ggtitle("Stacked Bars") +
  coord_flip()

p_sb2 <- ggplot(h, aes(x = genhz2, fill = texture_class)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  theme(aspect.ratio = 1) +
  ggtitle("Proportional Stacked Bars") + 
  coord_flip()

gridExtra::grid.arrange(p_sb1, p_sb2, ncol = 2)

```


---


## 3rd Dimension - Groups

```{r, echo=FALSE, fig.dim = c(8, 4)}

idx <- h$genhz %in% c("Cr", "R", "not-used")

p_b <- ggplot(h[!idx, ], aes(x = genhz2, y = clay)) + 
  geom_boxplot() +
  ylab("clay (%)") + xlab("genhz") +
  theme(aspect.ratio = 1) +
  coord_flip()

p_d <- ggplot(h[!idx, ], aes(x = clay, col = genhz)) +
  geom_density(alpha = 0.5, lwd = 1.5) +
  xlab("%") +
  theme(aspect.ratio = 1)

gridExtra::grid.arrange(p_b, p_d, ncol = 2)


```


---


## 3rd Dimension - Facets

```{r, echo = FALSE, fig.dim = c(8, 4)}

df <- reshape2::melt(h, 
                     id.vars = c("peiid", "genhz2", "hzdepm"), 
                     measure.vars = c("clay", "phfield", "total_frags_pct")
                     )

idx <- df$genhz2 %in% c("Cr", "R", "not-used")

ggplot(df[!idx, ], aes(y = value, x = genhz2)) +
  geom_boxplot() +
  xlab("genhz") +
  facet_wrap(~ variable, scales = "free_x") +
  coord_flip()
```


---


## 3rd Dimension - Facets

```{r, echo=FALSE, fig.dim=c(8, 4)}

s <- aqp::slice(loafercreek, 1:100 ~ clay + phfield + total_frags_pct)
s <- aqp::slab(s, fm = ~ clay + phfield + total_frags_pct, slab.fun = function(x) quantile(x, c(0.1, 0.5, 0.9), na.rm = TRUE))
names(s) <- gsub("\\.", "", names(s))
names(s) <- gsub("^X", "p", names(s))


ggplot(s, aes(x = top, y = p50)) +
  geom_line() +
  geom_ribbon(aes(ymin = p10, ymax = p90, x = top), alpha = 0.2) +
  xlim(c(100, 0)) + xlab("depth (cm)") + ylab("") +
  coord_flip() +
  facet_wrap(~ variable, scales = "free_x") +
  ggtitle("Loafercreek")

```


---


## 3rd Dimension - Correlation

```{r, echo=FALSE, fig.dim = c(8, 4)}

h <- transform(h, 
               frags = total_frags_pct,
               depth = hzdepm
               )
vars <- c("depth", "clay", "phfield", "frags")

GGally::ggpairs(h[vars])

```


---


## ggplot2 basics

```{r, eval=TRUE}

h <- horizons(loafercreek)

# set figure axis and aesthetics
p <- ggplot(data = h, aes(x = clay, y = hzdept, color = genhz))

# add layers
p + geom_point()

```


---

## Questions

1. What do the L, RV, and H represent?

2. What would cause the mean and median to be significantly different?

3. What are quantiles?

4. What is the difference between a histogram and bar plot?

5. How can you add a 3rd dimension to plots?


---


## Summary of Components

```{r, echo=FALSE, fig.dim = c(8, 4)}

co <- get_component_from_SDA(WHERE = "compname = 'Miami'")
co$slope_phase <- cut(co$slope_r, c(0, 2, 4, 8, 15, 30, 50))
co2 <- subset(co, majcompflag == "Yes")
h  <- get_chorizon_from_SDA(WHERE = "compname = 'Miami'")
test <- subset(h, cokey %in% co2$cokey)
test <- merge(test, co[c("cokey", "erocl")], by = "cokey", all.x = TRUE)

test2 <- test
depths(test2) <- cokey ~ hzdept_r + hzdepb_r
test3 <- aqp::slice(test2, seq(0, 100, 2) ~ claytotal_r + om_r + caco3_r + erocl)
test3 <- horizons(test3)
test3$hzdepm_r <- with(test3, (hzdepb_r + hzdept_r) / 2)

hvars <- c("claytotal_r", "om_r", "caco3_r")
pvars <- c("erocl")
test4 <- reshape(test3[c("cokey", "hzdept_r", hvars, pvars)],
                direction = "long",
                timevar = "variable", times = hvars,
                v.names = "value",    varying = hvars
                )
test4$variable <- factor(test4$variable, levels = c("claytotal_r", "om_r", "caco3_r"))
test4$erocl <- factor(test4$erocl, levels = levels(test4$erocl), labels = c("None - deposition", "Slightly", "Moderately", "Severely", "Extremely")[1:length(levels(test4$erocl))])


# ordering of groups is handled by ggplot2
ggplot(test4, aes(x = hzdept_r, y = value)) +
  geom_step(aes(group = cokey), direction = "vh", alpha = 0.1) +
  # geom_line(aes(group = cokey), alpha = 0.1) +
  geom_smooth(aes(col = erocl), se = FALSE, lwd = 2) +
  xlim(100, 0) +
  xlab("depth (cm)") + ylab("Percent (%)") +
  coord_flip() +
  facet_grid(~ variable, scale = "free_x") +
  labs(col = "Erosion Class") +
  ggtitle("Miami Components")

```


---


## Additional Reading

Healy, K., 2018. Data Visualization: a practical introduction. Princeton University Press. [http://socviz.co/](http://socviz.co/)

Helsel, D.R., and R.M. Hirsch, 2002. Statistical Methods in Water Resources Techniques of Water Resources Investigations, Book 4, chapter A3. U.S. Geological Survey. 522 pages. [http://pubs.usgs.gov/twri/twri4a3/](http://pubs.usgs.gov/twri/twri4a3/)

Kabacoff, R.I., 2015. R in Action. Manning Publications Co. Shelter Island, NY. [https://www.statmethods.net/](https://www.statmethods.net/)

Kabacoff, R.I., 2018. Data Visualization in R. [https://rkabacoff.github.io/datavis/](https://rkabacoff.github.io/datavis/)

Peng, R. D., 2016. Exploratory Data Analysis with R. Leanpub.  [https://bookdown.org/rdpeng/exdata/](https://bookdown.org/rdpeng/exdata/)  

Wilke, C.O., 2019. Fundamentals of Data Visualization. O'Reily Media, Inc. [https://serialmentor.com/dataviz/](https://serialmentor.com/dataviz/)
