---
title: Chapter 4 - Exploratory Data Analysis
author: Stephen Roecker and Tom D'Avello
date: "`r Sys.Date()`"
output:
  slidy_presentation:
    css: css/hpstr.css
    fig_caption: yes
    fig_height: 3
    fig_width: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(warning=FALSE, message=FALSE, results='asis')

library(soilDB)
library(ggplot2)

data(loafercreek)
data("metadata")

# Generalized the horizon designations
n <- c("A",
       "BAt",
       "Bt1",
       "Bt2",
       "Cr",
       "R")
# REGEX rules
p <- c("A",
       "BA|AB",
       "Bt|Bw",
       "Bt3|Bt4|2B|C",
       "Cr",
       "R")

loafercreek$genhz <- generalize.hz(loafercreek$hzname, n, p)

h <- horizons(loafercreek)
h <- within(h, {
  hzdepm = (hzdept + hzdepb) / 2
  texture_class = factor(texture_class, 
                         levels = metadata[metadata$ColumnPhysicalName == "texcl", "ChoiceName"]
                         )
  })


```


## Objectives

* Determine Low, RV, and High values (e.g. mean, median, etc...) for component and horizon data.
* Visualize your data to identify trends, using Rs base and ggplot2 graphics.
* Learn how to use the soilReports package to generate Rmarkdown reports.
* Learn how to properly transform your data for analysis.


---


##  Descriptive Statistics

```{r, echo=FALSE}

desc <- data.frame(
  Parameter = c("Mean", "Median", "Mode", "Standard Deviation", "Quantiles"),
  NASIS = c(rep("RV", 3), rep("L & H", 2)),
  Description = c("arithmetic average", 
                  "middle value, 50% quantile", 
                  "most frequent value", "variation", 
                  "percent rank of values, such that all values are <= p"
                  ),
  'R function' = c("mean()", "median()", "sort(table(), decreasting = TRUE)[1]", "sd()", "quantile()"),
  check.names = FALSE
  )
knitr::kable(desc)

```
 

---


## Graphical Methods - Descriptions

```{r, echo=FALSE}

figs <- data.frame(
  'Plot Types' = c("Bar", "Histogram", "Density", "Quantile-Quantile", "Box-Whisker", "Scatter & Line"),
   Description = c("a plot where each bar represents the frequency of observations for a 'group'",
                   "a plot where each bar represents the frequency of observations for a 'given range of values'",
                  "an estimation of the frequency distribution based on the sample data",
                  "a plot of the actual data values against a normal distribution",
                  "a visual representation of median, quartiles, symmetry, skewness, and outliers",
                   "a graphical display of one variable plotted on the x axis and another on the y axis"
                  ),
  check.names = FALSE
  )

knitr::kable(figs)

```


---
 

## Graphical Methods - Functions

```{r, echo=FALSE}

figs <- data.frame(
  'Plot Types' = c("Bar", "Histogram", "Density", "Quantile-Quantile", "Box-Whisker", "Scatter & Line"),
  # Description = c("a bar plot where each bar represents the frequency of observations for a given range of values",
  #                "an estimation of the frequency distribution based on the sample data",
  #                "a plot of the actual data values against a normal distribution",
  #                "a visual representation of median, quartiles, symmetry, skewness, and outliers",
  #                 "a graphical display of one variable plotted on the x axis and another on the y axis",
  #                "plots formatted for the representation of circular data"
  #                ),
  'Base R' = c("barplot()", "hist()", "plot(density())", "qqnorm()", "boxplot()", "plot()"),
  'lattice' = c("barchart()", "histogram()", "densityplot()", "qq()", "bwplot()", "xyplot"),
  'ggplot geoms' = c("geom_bar()", "geom_histogram()", "geom_density()", "geom_qq()", "geom_boxplot()", "geom_point()"),
  check.names = FALSE
  )

knitr::kable(figs)

```


---


## Distributions

```{r, echo=FALSE, fig.dim = c(8, 4)}

idx <- is.na(h$texture_class)

# bar plot
p_b <- ggplot(h[!idx, ], aes(x = texture_class)) +
  geom_bar() +
  ylab("count (n)") + xlab("texture class") +
  theme(aspect.ratio = 1) +
  ggtitle("Bar Plot")

# histogram
p <- ggplot(data = h, aes(x = clay))

p_h <- p + 
  geom_histogram() + 
  xlab("clay (%)") + ylab("count (n)") +
  theme(aspect.ratio = 1) +
  ggtitle("Histogram")

# density plot
p_d <- p + 
  geom_density(fill = "grey", alpha = 0.5) + 
  xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Density Plot")

gridExtra::grid.arrange(p_b, p_h, p_d, ncol = 3)

```


---


## The 'Normal Distribution

```{r, echo = FALSE, fig.dim = c(8, 4)}

dnorm_limit_1 <- function(x) {
  y <- dnorm(x)
  y[-1 > x | x > 1] <- NA
  return(y)
  }
dnorm_limit_2 <- function(x) {
  y <- dnorm(x)
  y[-2 > x | x > 2] <- NA
  return(y)
  }

# ggplot() with dummy data
ggplot(data.frame(x=c(-3, 3)), aes(x=x)) + 
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm_limit_1, geom="area", fill="blue", alpha=0.2) +
  stat_function(fun=dnorm_limit_2, geom="area", fill="orange", alpha=0.2) +
  annotate("text", x = 0, y = 0.2, label = "1 sd = 68%") +
  annotate("text", x = -1.5, y = 0.03, label = "2 sd = 95%")

```


---


## Mean vs Median


```{r, echo=FALSE, fig.dim = c(8, 4)}

p   <- c(0.025, 0.25, 0.5, 0.75, 0.975)

avg <- mean(h$clay, na.rm = TRUE)
std <- sd(h$clay,   na.rm = TRUE)

clay <- rbind(
  data.frame(
    value = c(avg - 2 * std, avg, avg + 2 * std),
    variable = "mean & sd",
    stringsAsFactors = TRUE
    ),
  data.frame(
    value    = quantile(h$clay, p, na.rm = TRUE),
    variable = "quantiles",
    stringsAsFactors = TRUE
    )
  )

avg <- mean(h$total_frags_pct, na.rm = TRUE)
std <- sd(h$total_frags_pct,   na.rm = TRUE)

frags <- rbind(
  data.frame(
    variable = "mean & sd",
    value = c(avg - 2 * std, avg, avg + 2 * std),
    stringsAsFactors = TRUE
    ),
  data.frame(
    variable = "quantiles", 
    value    = quantile(h$total_frags_pct, p, na.rm = TRUE),
    stringsAsFactors = TRUE
    )
  )

p1 <- ggplot(h, aes(x = clay)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(data = clay, aes(xintercept = value, lty = variable)) +
  xlab("%") +
  #xlim(-5, max(h$total_frags_pct)) +
  theme(aspect.ratio = 1) +
  ggtitle("Clay")

p2 <- ggplot(h, aes(x = total_frags_pct)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(data = frags, aes(xintercept = value, lty = variable)) +
  xlab("%") +
  #xlim(-5, max(h$total_frags_pct)) +
  theme(aspect.ratio = 1) +
  ggtitle("Rock Fragments")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```


---


## Distributions

```{r, echo=FALSE, fig.dim = c(8, 4)}

# qq plot
p_qq <- ggplot(h, aes(sample = clay)) + 
  geom_qq() +
  geom_qq_line() +
  xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Q-Q Plot")

# box plot
p_box <- ggplot(h, aes(x = 1, y = clay)) + 
  geom_boxplot() +
  ylab("clay (%)") +
  theme(aspect.ratio = 1) +
  coord_flip() +
  ggtitle("Box Plot")

gridExtra::grid.arrange(p_box, p_qq, ncol = 2)

```


---


## Relationships

```{r, echo=FALSE, fig.dim = c(8, 4)}

# scatter plot
p_s <- ggplot(h, aes(x = clay, y = hzdept)) +
  geom_point() +
  ylim(100, 0) +
  ylab("depth (cm)") + xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Scatter Plot")

# line graph
depth <- c(1, 5, 15, 30, 50, 70, 90)
lo_fit <- loess(clay ~ hzdepm, data = h, na.action = na.exclude)
clay <- predict(lo_fit, depth)
df <- data.frame(depth, clay)

p_l <- ggplot(h, aes(y = clay, x = hzdepm)) +
  geom_point(data = df, aes(x = depth, y = clay), size = 2, show.legend = FALSE) +
  geom_smooth(se = FALSE, col = "black") +
  xlim(100, 0) + ylim(min(h$clay, na.rm = TRUE), max(h$clay, na.rm = TRUE)) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1) +
  ggtitle("Line Plot")

gridExtra::grid.arrange(p_s, p_l, ncol = 2)

```


---


## 3rd Dimension - Color, Shape, Size, Layers, etc...

```{r, echo=FALSE, fig.dim = c(8, 4)}
h$clay2 <- ifelse(is.na(h$clay), 0, h$clay)

ggplot(h[!idx, ], aes(y = clay, x = hzdepm, col = genhz)) +
  geom_point(size = 2) +
  geom_smooth(se = FALSE, col = "black") +
  xlim(100, 0) + ylim(min(h$clay, na.rm = TRUE), max(h$clay, na.rm = TRUE)) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1)

```


---


## 3rd Dimension - Color

```{r, echo=FALSE, fig.dim = c(8, 4)}

h$genhz2 <- factor(h$genhz, levels = rev(levels(h$genhz)))

p_sb1 <- ggplot(h, aes(x = genhz2, fill = texture_class)) +
  geom_bar(position = "stack") +
  theme(aspect.ratio = 1) +
  ggtitle("Stacked Bars") +
  coord_flip()

p_sb2 <- ggplot(h, aes(x = genhz2, fill = texture_class)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  theme(aspect.ratio = 1) +
  ggtitle("Proportional Stacked Bars") + 
  coord_flip()

gridExtra::grid.arrange(p_sb1, p_sb2, ncol = 2)

```


---


## 3rd Dimension - Groups

```{r, echo=FALSE, fig.dim = c(8, 4)}

idx <- h$genhz %in% c("Cr", "R", "not-used")

p_b <- ggplot(h[!idx, ], aes(x = genhz2, y = clay)) + 
  geom_boxplot() +
  ylab("clay (%)") + xlab("genhz") +
  theme(aspect.ratio = 1) +
  coord_flip()

p_d <- ggplot(h[!idx, ], aes(x = clay, col = genhz)) +
  geom_density(alpha = 0.5, lwd = 1.5) +
  xlab("%") +
  theme(aspect.ratio = 1)

gridExtra::grid.arrange(p_b, p_d, ncol = 2)


```


---


## 3rd Dimension - Facets

```{r, echo = FALSE, fig.dim = c(8, 4)}

df <- reshape2::melt(h, 
                     id.vars = c("peiid", "genhz2", "hzdepm"), 
                     measure.vars = c("clay", "phfield", "total_frags_pct")
                     )

idx <- df$genhz2 %in% c("Cr", "R", "not-used")

ggplot(df[!idx, ], aes(y = value, x = genhz2)) +
  geom_boxplot() +
  xlab("genhz") +
  facet_wrap(~ variable, scales = "free_x") +
  coord_flip()
```


---


## 3rd Dimension - Correlation

```{r, echo=FALSE, fig.dim = c(8, 4)}

h <- transform(h, 
               frags = total_frags_pct,
               depth = hzdepm
               )
vars <- c("depth", "clay", "phfield", "frags")

GGally::ggpairs(h[vars])

```


---


## ggplot2 basics

```{r, eval=TRUE}

h <- horizons(loafercreek)

# set figure axis and aesthetics
p <- ggplot(data = h, aes(x = clay, y = hzdept, color = genhz))

# add layers
p + geom_point()

```


---


## Data Inspection - Missing Data

 1. **Exclude** all rows or columns that contain missing values using the function `na.exclude()`, such as `h2 <- na.exclude(h)`. However this can be wasteful because it removes all rows (e.g., horizons), regardless if the row only has 1 missing value. Instead it's sometimes best to create a temporary copy of the variable in question and then remove the missing variables, such as `clay <- na.exclude(h$clay)`.
 2.	**Replace** missing values with another value, such as zero, a global constant, or the mean or median value for that column, such as `h$clay <- ifelse(is.na(h$clay), 0, h$clay) # or h[is.na(h$clay), ] <- 0`.
 3. **Read** the help file for the function you're attempting to use. Many functions have additional arguments for dealing with missing values, such as `na.rm`.


---


## Additional Reading

Peng, R. D., 2016. Exploratory Data Analysis with R. Leanpub.  [https://bookdown.org/rdpeng/exdata/](https://bookdown.org/rdpeng/exdata/)  

Wickham, H., and G. Grolemund, 2017, 7 - Exploratory Data Analysis, pp: 81-109. In H. Wickham and G. Grolemund. R for Data Science. O'Reilly Media Inc., Sebastopol, CA. [http://r4ds.had.co.nz/exploratory-data-analysis.html](http://r4ds.had.co.nz/exploratory-data-analysis.html)

Healy, K., 2018. Data Visualization: a practical introduction. Princeton University Press. [http://socviz.co/](http://socviz.co/)

Helsel, D.R., and R.M. Hirsch, 2002. Statistical Methods in Water Resources Techniques of Water Resources Investigations, Book 4, chapter A3. U.S. Geological Survey. 522 pages. [http://pubs.usgs.gov/twri/twri4a3/](http://pubs.usgs.gov/twri/twri4a3/)
