# Zonal Statistics

To summarize spatial data for a polygon, some form of zonal statistics can be used. Zonal statistics is a generic term for statistics that aggregate data for an area or zone (e.g., a set of map unit polygons). This can be accomplished via two methods. The most common method provided by most GIS is the **census survey method**, which computes a statistical summary of all the raster cells that overlap a polygon or map unit. This approach is generally faster and provides a complete summary of the spatial data. An alternative approach is the **sample survey method**, which takes a collection of random samples from each polygon or map unit. While the sample approach is generally slower and does not sample every cell that overlaps a polygon it does offer certain advantages. For example the census approach used by most GIS typically only provides basic statistics such as: the min, max, mean, standard deviation, and sum. However, for skewed data sets the mean and standard deviation may be unreliable. A better alternative for skewed data sets is to use non-parametric statistics like quantiles. Examples of non-parametric statistics are covered in Chapter 4. The advantage of the sample approach is that it allows us to utilize alternative statistics, such as quantiles, and perform a more thorough exploratory analysis. While some people might prefer the census approach because it provides a complete summary of all the data that overlaps a map unit, it is important to remember that all spatial data are only approximations of the physical world and therefore are only estimates themselves with varying levels of precision.

Before extracting spatial data for the purpose of spatial prediction, it is necessary that the data meet the following conditions:  
  
- All data conform to a common projection and datum
- All raster data have a common cell resolution
- All raster data are co-registered, that is, the geographic coordinates of cell centers are the same for all layers. Setting the _Snap Raster_ in the ArcGIS Processing Environment prior to the creation of raster derivatives can ensure cell alignment. An ERDAS model is also available to perform this task.  

## Zonal Statistics in R

Zonal statistics in R can be implemented using either the census or sample approach. While R can compute zonal statistics using the census approach with the `zonal()` function in the raster package, it is considerably faster to call another GIS via the RSAGA or spgrass6 packages. These additional GIS packages provide R functions to access SAGA and GRASS commands. 

### Census Survey Example

For this example the RSAGA package will be used.

```{r r_poly_extract, eval = FALSE}
# load rgdal package
library(rgdal)

# import CA794 map unit polygons
ca794 <- readOGR(dsn = "E:/geodata/project_data/8VIC/ca794", layer = "ca794") 

# reproject
ca794 <- spTransform(ca794, CRS("+init=epsg:5070")) 

# convert MUKEY to  an integer
ca794$mukey2 <- as.integer(as.character(ca794$MUKEY)) 

# export shapefile
writeOGR(ca794, 
         dsn = "C:/workspace2", 
         layer = "ca794", 
         driver = "ESRI Shapefile", 
         overwrite_layer = TRUE
) 

# load RSAGA package
library(RSAGA)

# set rsaga path
myenv <- rsaga.env(path = "C:/Program Files/QGIS Essen/apps/saga") 

# load DEM
ned <- raster("E:/geodata/project_data/8VIC/sdat/ned30m_8VIC.sdat") 

# create a blank raster that matches the DEM
test <- raster(extent(ca794), ext = extent(ned), crs = crs(ned), res = res(ned))  

# export the raster
writeRaster(test, 
            file = "E:/geodata/project_data/8VIC/sdat/ca794.sdat", 
            format = "SAGA", 
            progress = "text", 
            overwrite = TRUE
) 

# Convert the CA794 shapefile to a rsaga raster
rsaga.geoprocessor("grid_gridding", 0, env = myenv, list(
  INPUT = "C:/workspace2/ca794.shp",
  FIELD = "mukey2",
  OUTPUT = "2",
  TARGET = "0",
  GRID_TYPE = "2",
  USER_GRID = "E:/geodata/project_data/8VIC/sdat/ca794.sgrd",
  USER_XMIN = extent(test)[1] + 15,
  USER_XMAX = extent(test)[2] - 15,
  USER_YMIN = extent(test)[3] + 15,
  USER_YMAX = extent(test)[4] - 15,
  USER_SIZE = res(test)[1]
))

# Extract the zonal statistics for 2 rasters
rsaga.geoprocessor("statistics_grid", 5, env = myenv, list(
  ZONES = "E:/geodata/project_data/8VIC/sdat/ca794.sgrd",
  STATLIST = paste(c("E:/geodata/project_data/8VIC/sdat/ned30m_8VIC.sgrd", "E:/geodata/project_data/8VIC/sdat/ned30m_8VIC_slope5.sgrd"), collapse = ";"),
  OUTTAB = "C:/workspace2/github/stats_for_soil_survey/trunk/data/ca794_zonal.csv"
))
```
```{r poly extract2, eval=FALSE}
# import the results
test <- read.csv("C:/workspace2/github/ncss-tech/stats_for_soil_survey/data/ca794_zonal.csv") 

# rename the mukey column
names(test)[1] <- "mukey" 

# examine mukey 2480977
subset(test, mukey == 2480977) 
```

### Sample Survey Example

To implement the sample approach to zonal statistics we can use the `extract()` and `over()` functions in the raster and sp packages respectively.

```{r, eval = FALSE}
# Take a stratified random sample
s <- spsample(ca794, n = 1000, type = "stratified")

setwd("E:/geodata/project_data/8VIC/ca794")

# Create a raster stack
rs <- stack(c(
  elev = "ned30m_8VIC.tif",
  slope = "ned30m_8VIC_slope5.tif")
)
# Set the spatial projection
proj4string(rs) <- CRS("+init=epsg:5070")

# Extract the map unit polygon value
test1 <- over(s, ca794)

# Extract the raster values
test2 <- data.frame(extract(rs, s))

# Combine the two datasets
test2 <- cbind(test1, test2)

# Cache/save the results
save(test2, file = "C:/workspace2/github/stats_for_soil_survey/trunk/data/ch2_sample.Rdata")
```

```{r, eval=FALSE}
# Load the cache/saved results
load(file = "C:/workspace2/github/ncss-tech/stats_for_soil_survey/data/ch2_sample.Rdata")

# Examine summary for mukey 2480977
summary(subset(test2, MUKEY == 2480977)) 
```

Compare the results of the census and sample approaches above. While the census approach surveyed 204,460 cells, the sample approach only surveyed 5,777. However we can see that the results are largely similar between the two approaches.


### Exercise 3: calculating zonal statistics

- Use your own data
- Extract the zonal statistics for one soil survey area or several polygons, using the sample approach
- Show your work and submit the results to your coach

