---
title: "Numerical Taxonomy Assignment 1"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Objectives
Using the concepts (and code) from the [Numerical Taxonomy and Ordination](http://ncss-tech.github.io/stats_for_soil_survey/book2/numerical-tax.html) chapter, compute pair-wise distances between soil series concepts from a suite of 8 physical and climatic variables. These 8 variables (*characteristics*) will be used to develop a climate signature for each soil series (*individual*). Present and interpret the results using both a dendrogram and ordination.

Key steps:

 * select your favorite soil series (5-10) and get the data via `fetchOSD`
 * develop the **data matrix**
 * develop the **distance matrix**, ideally after standardization of characteristics
 * develop a dendrogram via hierarchical cluster analysis
 * develop an ordination via `sammon` or `metaMDS`
 * interpret the results


## Brief Description of the Data Source
Annual and monthly climate summaries have been estimated from the [SSR2 standard stack of 1981--2010 PRISM data](https://ncss-tech.github.io/soilReports/docs/region2_mu-comparison-data-sources.html). [Weighted percentiles](https://www.rdocumentation.org/packages/Hmisc/versions/4.1-1/topics/wtd.stats) were estimated from a single sample from within each map unit delineation, weighted by 

$$log(polygon\ area * component\ percentage)$$

We will be using the annual climate summaries. See this [document](http://ncss-tech.github.io/AQP/soilDB/soil-series-query-functions.html) for more information on how summaries like these have been developed and how to access them via `fetchOSD`.


## Setup

First, you will need to load some packages. If you have completed the [pre-class assignment](http://ncss-tech.github.io/stats_for_soil_survey/chapters/0_pre-class-assignment/pre-class-assignment.html) then all of these packages should be in place.
```{r eval=TRUE}
library(soilDB)
library(sharpshootR)
library(latticeExtra)
library(reshape2)
library(RColorBrewer)
library(cluster)
library(ape)
library(vegan)
library(MASS)
```



# An Example

The `fetchOSD` function from the `soilDB` package is a simple interface data that have been extracted from the text OSDs and summarized from the current SSURGO snapshot. The `extended=TRUE` argument is used to access the climate data we will be using in this assignment.
```{r fig.width=10, fig.height=6}
soils <- c('Ava', 'Drummer', 'Cisne', 'Pierre', 'Cecil', 'Appling', 'San Joaquin', 'Redding', 'Corning')
s <- fetchOSD(soils, extended = TRUE)
```

Select percentiles are provided for 7 annual climate summaries and elevation. Note that "q50" is the median value and "n" is the number of samples (i.e. number of delineations) used to estimate the percentiles.
```{r echo=FALSE}
kable_styling(kable(s$climate.annual[1:8, ], format='html', digits = 2), font_size = 9, full_width = FALSE)
```


The data look something like this. Your assignment is to compute pair-wise distances between series concepts, using median values (filled circles in the figure) from these data.
```{r fig.width=12, fig.height=5.5, echo=FALSE}
segplot(factor(series) ~ q05 + q95 | climate_var, centers=q50, data=s$climate.annual, draw.bands=FALSE, segments.fun=panel.arrows, ends='both', angle=90, length=1, unit='mm', scales=list(y=list(alternating=3), x=list(relation='free')), as.table=TRUE, col='RoyalBlue', strip=strip.custom(bg=grey(0.85), par.strip.text=list(cex=0.7)), xlab='5th-25th-50th-75th-95th Percentiles', panel=function(x, y, z, q25=s$climate.annual$q25, q75=s$climate.annual$q75, subscripts, ...) {
  # basic plot
  panel.grid(h=FALSE, v=-1, col='grey', lty=3)
  panel.abline(h=1:length(unique(s$climate.annual$series)), col='grey', lty=3)
  panel.segplot(x, y, z, subscripts=subscripts, ...)
  
  # add interquartile range
  q25 <- q25[subscripts]
  q75 <- q75[subscripts]
  zz <- z[subscripts]
  panel.rect(xleft=q25, xright=q75, ybottom=as.numeric(zz) - 0.025, ytop=as.numeric(zz) + 0.025, border='RoyalBlue', col='RoyalBlue')
})
```

The results might look something like this.
```{r echo=FALSE, fig.width=8.5, fig.height=4, results='hide'}
x <- s$climate.annual
x.wide <- dcast(x, series ~ climate_var, value.var = 'q50')
row.names(x.wide) <- x.wide$series

d <- daisy(x.wide[, -1], stand = TRUE, metric = 'gower')
dd <- diana(d)
h <- as.hclust(dd)
p <- as.phylo(h)

# define colors based on cutting a divisive hierarchical clustering into 4 groups
cols <- brewer.pal(9, 'Set1')[cutree(h, 4)]

# nMDS from distance matrix
s <- sammon(d)

par(mar=c(1,1,2,1), mfcol=c(1,2))
plot(p, label.offset=0.125, direction='right', font=1, cex=0.85, main='Divisive Hierarchical Clustering')
tiplabels(pch=15, col=cols)


plot(s$points, asp=1, type='n', axes=FALSE, xlab='', ylab='', main="nMDS by Sammon's Non-Linear Mapping", xlim=c(-0.3, 0.6))
abline(v=0, h=0, col='black')
grid()
text(s$points, rownames(s$points), cex=0.65, col=cols, font=2)
box()
```

---------------------


# Your Turn
We will help with the data preparation but it will be up to you to finish the process.

## Get and Prepare Data
Adjust the vector of soil series names assigned to the `soils` character vector below. 5 to 10 should be enough, but feel free to use all [50 state soils](https://raw.githubusercontent.com/ncss-tech/soil-color-book/master/state-soils/state_soils.csv).
```{r eval=FALSE}
# define series names and get data
## you will need to edit this accordingly
soils <- c('your favorite here', 'another favorite', 'maybe the state soil', ...)
s <- fetchOSD(soils, extended = TRUE)

# extract annual climate + elevation data
x <- s$climate.annual

# check structure of the data
str(x)
head(x)
```

Develop the **data matrix**. The data are in "long format" (multiple rows per series, one for each variable) but we need the data in "wide format" (single row per series, columns containing medians).
```{r eval=FALSE}
# re-shape into wide format
x.wide <- dcast(x, series ~ climate_var, value.var = 'q50')
```

Check the structure of `x.wide` one more time: note that the first column contains the soil series name. This is something we want to "keep track of" but not include in the pair-wise distance calculation.
```{r eval=FALSE}
head(x.wide)

# save the series names into row names for later
# remember code examples from chapter 5?
row.names(x.wide) <- x.wide$series


# consider two possible ways of excluding the first column
# [ ] subsetting via negative column index
# this doesn't change x.wide but is handy for an on-the-fly modification
head(x.wide[, -1])
```

## Develop Distance Matrix
Don't forget to standardize characteristics: `dist()` doesn't know how to standardize but `daisy()` does. See `?daisy` for clues.

```{r}
# dist() or daisy() 

```

## Hierarchical Clustering
There are several methods and linkage criteria. How do you choose?
```{r}
# hclust(), agnes(), diana()

# possibly convert to ape class for better figures, via hclust class
# as.hclust()
# as.phylo()
```

## Ordination
Note that this is based on the distance matrix. Be sure to inspect the structure of the results with `str()`.
```{r}
# sammon() or metaMDS()

```

## Visualize Pair-Wise Distances via Dendrogram and Ordination
```{r}
# most hierarchical clustering objects have a plot() method

```

## Interpret
 * Do these figures tell you anything that you didn't already know? 
 * Does the hierarchy suggested by the dendrogram mean anything? 
 * Does cutting the dendrogram into $n$ clusters result in groups that follow intuition? 
 * Which representation (dendrogram vs. ordination) is more useful? 
 * Do the axes of the 2D representation (e.g. ordination) of the original 8D data space map to meaningful climatic gradients? 
 * How might we use the pair-wise distances computed from select characteristics of all soil series concepts?




