---
title: "Clustering of Image Colors"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

## Objectives

In this example we will apply ideas from the [Numerical Taxonomy / Ordination](http://ncss-tech.github.io/stats_for_soil_survey/book2/numerical-tax.html) module to a photograph. These methods can also be applied to a stack of terrain/climate derivatives for the purposes of initial or update soil survey work. The basic work-flow looks something like this:

  * load stack of conformal grids: same coordinate reference system, grid size, grid extent
  * exhaustive sampling of grids (random, regular, etc.) at points
  * sub-sample points via cLHS
  * compute distance matrix
  * apply clustering algorithms
  * visualize: dendrogram (if possible), nMDS, etc.
  * fit simple model (LDA, etc.) to predict cluster ID at every pixel in the stack
  * make predictions, could be class membership or most-likely class
  * visualize predictions as an image
  * evaluate class-wise uncertainty with Shannon entropy



Several parts of this demonstration depend on functions that have a non-deterministic outcome: 

  * cLHS-sampling via `clhs::clhs()`
  * nMDS via `MASS::sammon()`
  * partitioning around medoids clustering via `cluster::pam()`
  
In order to create repeatable results, the following examples use the `set.seed()` function to control randomness. This can be a useful trick when it is neccessary to generate the same output each time a report or analysis is run. However, one should not use `set.seed()` when a truly random outcome is required.

## Setup

Load packages that we will be using.
```{r}
library(aqp)
library(terra)
library(cluster)
library(ape)
library(farver)
library(MASS)
library(colorspace)
library(viridisLite)
library(clhs)
```

Use an image with well-defined colors to demonstrate methods / syntax.
```{r}
# I like airplanes
# URL to example image, you can swap out for any image you like
url <- 'https://github.com/ncss-tech/stats_for_soil_survey/raw/master/exercises/numerical-taxonomy/photo.jpg'
tf <- tempfile(fileext = '.jpg')

# download, this may fail if you are on the VPN
download.file(url, destfile = tf, mode = 'wb')

# load as 3-band SpatRast object
# bands: red, green, blue (sRGB color space)
r <- rast(tf)

# optionally, use your own image or stack of GIS data
# r <- rast(...)
```

Inspect the resulting `SpatRast` object.
```{r}
# re-name bands for access later
names(r) <- c('r', 'g', 'b')

# object properties
print(r)
```


Display 3-band stack as RGB image.
```{r dev = 'jpeg'}
plotRGB(r)
```


## Color Conversion

Convert to [CIELAB](https://en.wikipedia.org/wiki/CIELAB_color_space) color space, much more useful for analysis of color. I'll abbreviate CIELAB as LAB for simplicity.
```{r dev = 'jpeg'}
# copy SpatRast / update with LAB color coordinates
r.lab <- r

# simple trick to directly access / replace values of a SpatRast
values(r.lab) <- convert_colour(values(r), from = 'rgb', to = 'lab', white_from = 'D65')
names(r.lab) <- c('l', 'a', 'b')

# check
# bands are L, A, B color space coordinates
# positive "A" coordinates are "red"
plot(r.lab, col = mako(25), axes = FALSE)
```


## Sampling

Exhaustive sampling, extracting LAB color coordinates.
```{r}
s <- spatSample(r.lab, size = 5000, method = 'regular', as.points = TRUE)
```

Sub-sample using cLHS algorithm. Much easier to compute distance matrix from a small, representative sample.
```{r}
# note temporary conversion to sf object
# required by clhs()
# result is an integer index to samples
set.seed(42)
idx <- clhs(sf::st_as_sf(s), size = 250)
```

Visual check on cLHS-selected samples (blue).
```{r dev = 'jpeg'}
plotRGB(r, axes = FALSE)
# points(s, pch = 3, col = 'white', cex = 0.25)
points(s[idx, ], pch = 1, col = 'blue', cex = 1.5)
```

Visualize LAB data.
```{r fig.width=7, fig.height=7}
# keep only the cLHS samples
s <- s[idx, ]

# convert SpatVect -> data.frame
s.lab <- as.data.frame(s)

## convert CIE LAB values extracted at smples -> sRGB
s.rgb <- convert_colour(s.lab[, c('l', 'a', 'b')], from = 'lab', to = 'rgb', white_from = 'D65')

# convert to R colors, hex notation
s.cols <- rgb(s.rgb, maxColorValue = 255)

# 3-variable scatter plot matrix
# CIELAB color space coordinates
# colors are from sRGB color coordinates
pairs(s.lab[, c('l', 'a', 'b')], pch = 16, col = s.cols)
```



## Distance Matrix

Compute pair-wise distances via $\Delta{E_{00}}$ [color contrast metric](https://en.wikipedia.org/wiki/Color_difference).
```{r}
## compute pair-wise distances
# using the CIE2000 color contrast metric
# perceptual (avg. human vision) vs. Euclidean distance
d <- compare_colour(
  from = s.lab[, c('l', 'a', 'b')], 
  to = s.lab[, c('l', 'a', 'b')], 
  from_space = 'lab', 
  method = 'CIE2000'
)

# convert full-matrix form to simpler 'dist' representation
d <- as.dist(d)

# check for 0 distances, can cause problems for nMDS
summary(d)

# if there are any, set those to a small value
d[which(d == 0)] <- quantile(d[which(d > 0)], prob = 0.001, na.rm = TRUE)
```

## Ordination by nMDS
```{r fig.width=7, fig.height=7}
# "fixed" randomness, 
# only important when you want the same result every time
set.seed(42)
mds <- sammon(d)
str(mds)

# simple visualization of nMDS ordination
# colors from sRGB color coordinates, converted to hex notation
par(mar = c(2, 2, 1, 1))
plot(mds$points, pch = 16, col = s.cols, cex = 2, axes = FALSE)
# add horizontal / vertical lines to see "origin" of nMDS coordinates
abline(h = 0, v = 0, lty = 3)
box()
mtext('nMDS Axis 1', side = 1, line = 0.5, font = 2)
mtext('nMDS Axis 2', side = 2, line = 0.5, font = 2)
```



## Cluster Analysis

### Divisive Hierarchical Clustering
Note that the `hclust` object is often used as a common format when converting from one object class to another. For example, converting the divisive hierarchical clustering object created by `diana()` to `phylo` class requires a trip through the `hclust` format.

```{r}
# divisive hierarchical clustering of distance matrix
dd <- diana(d)

# convert cluster pkg object -> hclust obj. -> phylo obj.
h <- as.phylo(as.hclust(dd))

# attempt rotation of the dendrogram leaves
# using rank-order from the first nMDS axis above
h <- rotateConstr(h, constraint = order(mds$points[, 1]))
```

Using the `plot` method for `phylo` class objects, as provided by the ape package.
```{r fig.width = 9, fig.height = 3.5, dpi= 100}
par(mar = c(0, 0, 2, 0))

plot(
  h, 
  show.tip.label = FALSE, 
  direct = 'down', 
  type = 'phylogram', 
  y.lim = c(0, 50),
  main = sprintf("Divisive Coef: %s", round(dd$dc, 2))
)

# make a block of tip label colors
# using repeated calls, varying vertical offsets
tiplabels(pch = 15, col = s.cols, offset = 1.5)
tiplabels(pch = 15, col = s.cols, offset = 2.5)
tiplabels(pch = 15, col = s.cols, offset = 3.5)
```

### K-medoids Clustering

```{r fig.width=6, fig.height=6}
# partitioning around medoids
# using the distance matrix vs. data matrix (diss = TRUE)
set.seed(42)
cl <- pam(d, k = 4, diss = TRUE)

# save clustering vector, these are integers
# interpret as categories
s.lab$cl <- factor(cl$clustering)

# plot method for partioning-style clustering (cluster package)
clusplot(cl, col.p = s.cols, plotchar = FALSE, col.clus = 'black')
```

Extract medoid colors in sRGB color space (for viewing on the screen).
```{r fig.width=6, fig.height=2}
# cl$medoids is an index to the original data matrix
# will use these later
medoid.cols <- rgb(s.rgb[cl$medoids, ], maxColorValue = 255)

# quick check on medoid colors
par(mar = c(0, 0, 0, 0))
swatchplot(medoid.cols)
```


## Supervised Classification
Supervised classification to interpolate between point samples, using linear discriminant analysis.
```{r}
# linear discriminant analysis (LDA) model works well here
# simple and efficient
m <- lda(cl ~ l + a + b, data = s.lab)
```

Demonstrate prediction output from `lda` object using some of the training data. The result is a `list`:

  * `class` is the most likely class, as a factor
  * `posterior` are the class-wise probabilities, as a matrix
  * `x` are the LDA "scores" (not used here)

```{r}
# use only the first 6 rows of data
p <- predict(m, newdata = s.lab[1:6, ])

p$class

round(p$posterior, 3)
```

Linear discriminant function scatter plot matrix.
```{r fig.width=6, fig.height=6}
plot(m, col = s.cols)
```

Wrapper functions around `predict` method used by LDA models, as applied to multi-band `spatRaster` objects.
```{r}
# prediction for most-likely class
pred.class <- function(model, data) {
  predict(model, data)$class
}

# class probabilities
pred.pr <- function(model, data) {
  predict(model, data)$posterior
}
```

Predict over entire stack (3-band) of LAB color coordinates.
```{r}
# most-likely cluster
p.class <- predict(r.lab, m, fun = pred.class)

# cluster probabilities
p.pr <- predict(r.lab, m, fun = pred.pr)
```



## Cluster Images (Maps)

Most-likely cluster, for each pixel.
```{r dev = 'jpeg'}
plot(p.class, col = medoid.cols, axes = FALSE)
```

Cluster probabilities.
```{r}
par(mfrow = c(2, 2))

for(i in 1:4) {
  
  # smooth color ramp
  # from white -> medoid color i
  cr <- colorRampPalette(c('white', medoid.cols[i]), space = 'Lab')
  
  plot(
    p.pr[[i]], 
    col = cr(10), 
    axes = FALSE, 
    main = sprintf("Cluster %s", i)
  )
}
```

Shannon entropy.
```{r dev = 'jpeg'}
# apply aqp::shannonEntropy() function at each stack of pixels
# stack represents predicted class-wise probability
H <- app(p.pr, shannonEntropy)

plot(H, col = viridis(25), axes = FALSE, main = 'Shannon Entropy (log base 2)')
```


## nMDS Images (Maps)
Using multiple linear regression, we can estimate the nMDS scores for all pixels in the 3-band image.
```{r}
# row-order is preserved
# transfer nMDS scores to source data
# 250 sub-samples
s.lab$mds1 <- mds$points[, 1]
s.lab$mds2 <- mds$points[, 2]

# multiple linear regression of nMDS scores
# via CIE LAB color coordinates
lm.mds1 <- lm(mds1 ~ l + a + b, data = s.lab)
lm.mds2 <- lm(mds2 ~ l + a + b, data = s.lab)

# internal validation metrics, looks OK
summary(lm.mds1)
summary(lm.mds2)

# spatial prediction
p.mds1 <- predict(r.lab, lm.mds1)
p.mds2 <- predict(r.lab, lm.mds2)
```

Attempt a simple interpretation of nMDS scores (axes).
```{r fig.width=7.5, fig.height=4, dev = 'jpeg'}
# colors
mds1.cols <- colorRampPalette(c('red', 'green'), space = 'Lab')(25)
mds2.cols <- rev(grey.colors(25))

# maps
par(mfcol = c(1, 2))
plot(p.mds1, col = mds1.cols, axes = FALSE, main = 'nMDS Axis 1', mar = c(2, 1, 2, 3))
mtext(text = '"red \U2192 green"', side = 1, line = 0.5)

plot(p.mds2, col = mds2.cols, axes = FALSE, main = 'nMDS Axis 2', mar = c(2, 1, 2, 3))
mtext(text = '"light \U2192 dark"', side = 1, line = 0.5)
```





