---
title: "nMDS Evaluation"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Introduction
Compare the ordination by nMDS of two simulated data sets using scree plots of "stress" and the Shepard Diagram. Data matrix `X1` represents high-rank (10 dimensions) noise, a poor candidate for ordination in 2 dimensions. Data matrix `X2` represents lower rank (4 dimensions) data with some natural clustering, a good candidate for ordination in 2 dimensions.


Simulate data.
```{r results='hide'}
library(MASS)

# number of individuals 
.n <- 100

# completely unstructured, 10-dimensional data = noise
set.seed(42)
x1 <- matrix(rnorm(.n * 10, mean = 0, sd = 0.1), ncol = 10)

# data with some structure, 4-dimensions, still noisy
set.seed(42)
x2 <- rbind(
  matrix(rnorm(.n, mean = 0, sd = 1), ncol = 4),
  matrix(rnorm(.n, mean = 0.1, sd = 1), ncol = 4),
  matrix(rnorm(.n, mean = 0.2, sd = 1), ncol = 4),
  matrix(rnorm(.n, mean = 2, sd = 0.5), ncol = 4)
)


# Euclidean distance
d1 <- dist(x1)
d2 <- dist(x2)


# evaluate stress as function of ordination rank
.seq <- 1:6

.stress1 <- sapply(.seq, function(i) {
  .res <- sammon(d1, k = i)$stress
})

.stress2 <- sapply(.seq, function(i) {
  .res <- sammon(d2, k = i)$stress
})
```

Scree plots of stress. Three dimensions seems like a reasonable target.
```{r fig.width = 5, fig.height = 5}
plot(x = .seq, y = .stress1, type = "b", las = 1, lwd = 2, ylab = "Stress", xlab = 'Number of Dimensions', main = "Sammon's Non-Linear Mapping (nMDS)", ylim = c(0, max(c(.stress1, .stress2))))

lines(x = .seq, y = .stress2, type = 'b', col = 2, lwd = 2)

legend('topright', legend = c('X1', 'X2'), lwd = 2, col = 1:2, bty = 'n')
```

Attempt nMDS along 3 axes (dimensions).
```{r results='hide'}
# nMDS
set.seed(42)
mds1 <- sammon(d1, k = 3)

set.seed(42)
mds2 <- sammon(d2, k = 3)

# generate data for Shepard Diagram
shep1 <- Shepard(d1, mds1$points) 
shep2 <- Shepard(d2, mds2$points) 
```

nMDS scores, first two axes.
```{r fig.width = 8.5, fig.height = 5}
par(mfcol = c(1, 2))
plot(mds1$points[, 1:2], pch = 16, asp = 1, xlab = 'nMDS Axis 1', ylab = 'nMDS Axis 2', main = 'X1', las = 1, cex = 0.66)

plot(mds2$points[, 1:2], pch = 16, asp = 1, xlab = 'nMDS Axis 1', ylab = 'nMDS Axis 2', main = 'X2', las = 1, cex = 0.66)
```

The poor nMDS solution for `X1` is clearly indicated by the general lack of fit in the Shepard diagrams. In other words, the pair-wise distances within the reduced set of ordination axes do not reliably describe the pair-wise distances of the original data-space.
```{r fig.width = 8.5, fig.height = 5}
par(mfcol = c(1, 2))
plot(shep1, pch = ".", asp = 1, xlab = 'Original Distance', ylab = 'nMDS Distance', main = 'X1: Shepard Diagram', las = 1)
lines(shep1$x, shep1$yf, type = "S", col = 2, lwd = 2)
abline(0, 1, col = 4, lwd = 2, lty = 2)

plot(shep2, pch = ".", asp = 1, xlab = 'Original Distance', ylab = 'nMDS Distance', main = 'X2: Shepard Diagram', las = 1)
lines(shep2$x, shep2$yf, type = "S", col = 2, lwd = 2)
abline(0, 1, col = 4, lwd = 2, lty = 2)
```


