---
title: "Getting started"
author: "Wills"
date: "December 28, 2018"
output: html_document
---
## R Markdown

#### Setup
Include packages that will be used and options that apply to the entire document
__Note that adding include = False will cause the commands in the chunk to not be shown__
__{r setup, include=FALSE}__

```{r setup}

#if you've used R before - we'll clear your session
rm (list = ls(all = TRUE))

#if you need to unload packages use the following command
#lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)


library(knitr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)

#this will tell you information about your current R session 
##important for troubleshooting and when asking for help

sessionInfo()


```

## Data

I include a seperate chunk for data. It's not required, so you could do this as part of the setup.

```{r data}

sand2 <- read.table("C:/R_data/sand_example.csv", header=TRUE, sep=",", 
  na.strings="NA", dec=".", strip.white=TRUE)

```

## Base Plots
These are 'base' plots that can be created in R.  No extra packages are neeeded to run these.

```{r histograms}

plot(depth~sand, data = sand2, ylim =c(50,-1), main = "Total Sand (%) by depth", 
     xlab = "Sand", ylab = "Depth")

hist(sand2$sand)

#change color and labels

hist(sand2$sand, col = 'blue', main = 'Histogram of Sand Content', xlab = 'Sand Content (%)')


```

##An Alternate Plot Package:ggplot

These graphs require the ggplot2 package (it must be loaded with the libarary command within a code chunk.)  I avoided naming the chunk ggplot to avoide the confusion with the package. The # still comments lines in the code chunks.  Alternatively, you could put the comments into the text portion of the R markdown document.


```{r GG_PLOTS}
#simple histogram
ggplot(data = sand2, aes(x = sand )) + geom_histogram()

#we can control the number of bins
ggplot(data = sand2, aes(x = sand )) + geom_histogram(bins = 10)

#or the width of bins
ggplot(data = sand2, aes(x = sand )) + geom_histogram(binwidth = 1)

#we can also change the color
ggplot(data = sand2, aes(x = sand )) + geom_histogram(binwidth = 1, fill = 'blue')


#let's create a boxplot with ggplot
l <- ggplot(data = sand2, aes(x = landuse, y = sand)) + geom_boxplot()

#wait no plot showed up - that's because we assigned the plot the object 'l'

l

#we can add color by assigning fill to a variable
lc <- b <- ggplot(data = sand2, aes(x = landuse, y = sand)) + geom_boxplot(aes(fill = landuse))
lc

#we can make a more complicated graph by filling with another variable
lm <- b <- ggplot(data = sand2, aes(x = landuse, y = sand)) + geom_boxplot(aes(fill = master))
lm

#we can add new command to the exsisting object 
#note that was are using fill because that is what's in the aes(), sometimes you'll use scale_color_
lm + scale_fill_brewer(palette = "Set1")
lm + scale_fill_grey()


#Notice that there is a + as you build elements - you can have a return after a comma within a () without a +
#use R studio to help you track and find pairs of '' and ()
#and we can add labels and adjust ticks
lm + scale_fill_grey() + labs(x = "Landuse", y = "Sand Content (%)") +
  theme(axis.text.x = element_text(face="bold", color="#993333", size=14, angle=45),
        axis.text.y = element_text(face="bold", color="#993333", size=14, angle=45))

#our additions are necessarily better..........
#there are many ways edit graphs for presentations and sharing
#there are many helpful websites




```


Note that the `echo = FALSE` parameter CAN BE added to the code chunk to prevent printing of the R code that generated the plot.
