---
title: "Spatial Introduction Exercise Instructions"
author: "Andrew G. Brown"
date: "2/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Intro Instructions

This exercise will demonstrate how to use R Markdown to make interactive maps inside HTML reports. 

0. Install the development version of `mapview` using `remotes`.

```{r, eval=FALSE}
remotes::install_github("rspatial/mapview")
```

1. Open a new R Markdown document: _File >> New File >> R Markdown..._

2. Enter a **Title** and select `HTML` as the output format.

3. _Delete_ all text below the YAML header (lines 7 to 30). The YAML header is denoted by `---` and must begin on the first line. 

4. Enter the following code from Chapter 4, Exercise 1 (4.4.2)

 - http://ncss-tech.github.io/stats_for_soil_survey/book/spatial.html#exercise-1
 
````
```{r, eval=TRUE}`r ''`
library(aqp)
library(soilDB)
library(sf)
library(mapview)

# get pedons from the selected set
pedons <- fetchNASIS(from = 'pedons')

# REMOVE PEDONS WITH NA COORDINATES
pedons.sp <- subset(pedons, !is.na(x_std) & !is.na(y_std))

# create sf object (more on this in next section)
pedon.locations <- st_as_sf(
  site(pedons.sp), 
  coords = c('x_std','y_std'),
  crs = st_crs(4326) #WGS84 GCS
)

# plot an interactive map
mapview(pedon.locations, 
        legend = FALSE, 
        map.types = 'OpenStreetMap', 
        label = pedon.locations$site_id)
```
````

5. Click the Knit button (Yarn icon). You will be prompted to save your file if you have not already. Save it as `my-pedons_YOURNAME.Rmd`. The file name must end in _.Rmd_.

6. Inspect the output. Look at the messages that are produced from loading packages, and the interactive plot should be at the bottom of the page. 

 - What options are available on the map?
 
 - What happens when you click on a point?
