---
output: html_document
---
# Dichotomous Soil Sorting Key in R
## Greg Schmidtt

This strategy is based on known relationships of ecology to soil properties. 

It forces you to use the properties in the `SoilProfileCollection` to make a decision for every component.

I have seen many vague keys out there which only loosely translate into which component should be included, especially keys that are not dichotomous. The keys written in R can get complicated, but like taxonomic keys, you can break it into smaller units. 

Here’s an example wherein I’ve determined that a set of soils would meet the criteria for being “Sandy”, and then need to be broken down into groups based on hydrology:

The "Sandy" group was determined using the following key.

All sites have the label “Not” as their site name, and as they are processed by the key, categories that come out in earlier keys such as flood-plains get their own label so has to not be included in the subsequent keys. 

This also shows that I had to make some allowance for differences based on geography, which meant that I had to do some MLRA/LRU assignment in GIS and link to the data frames in R:

```{r}
library(aqp)
library(soilDB)
```

```{r, echo=FALSE, include=FALSE}
# inspection to find a suitable SDA dataset
# res <- SDA_query("SELECT DISTINCT ecoclassname, cokey FROM coecoclass 
#                   WHERE ecoclassname LIKE '%sandy depression%'")
#       
```

```{r}
s <- get_component_from_SDA("ecoclassname LIKE '%depression%'")
x <- get_chorizon_from_SDA(paste0("c.cokey IN ", format_SQL_in_statement(s$cokey)))

depths(x) <- cokey ~ hzdept_r + hzdepb_r
site(x) <- s
x <- fillHzGaps(x, to_bottom = 200)
```

Calculate a few properties at a few depths

```{r}
property_columns <- c("sandtotal_r", "claytotal_r", "ph1to1h2o_r")
subset_columns <- c(idname(x), property_columns)

s50 <- horizons(trunc(x, 49, 50)[,1])[, subset_columns]
s150 <- horizons(trunc(x, 149, 150)[,1])[, subset_columns]
colnames(s50) <- c(idname(x), paste0("T50_", c("sand", "clay", "pH")))
colnames(s150) <- c(idname(x), paste0("T150_", c("sand", "clay", "pH")))

# add carbonate depth (shallowest k subscript)
x$ksuffix <- grepl("k", x$hzname)
x$highcaco3 <- !is.na(x$caco3_r) & x$caco3_r > 10 
x$is_calcareous <- ifelse(x$ksuffix | x$highcaco3, "calcareous", "non-calcareous")
x$carbdepth <- minDepthOf(x, 
                          pattern = "^calcareous",
                          hzdesgn = "is_calcareous", 
                          no.contact.assigned = 200)$hzdept_r

# join in new site data (slices from target depths, properties renamed)
site(x) <- s50
site(x) <- s150

# remove components with NA data in key properties
x <- subset(x, complete.cases(site(x)[,unique(c("carbdepth", colnames(s50), colnames(s150)))]))
```

### Grouping of "Sandy" "Acid" and "Calcareous" sites

All components with an ecosite that has `"Sandy"` in the correlated ecological site name are assigned to the `"Sandy"` group. In addition, `"Sandy"` contains any component with 50cm sand >80%, or 50cm sand >70% and 150cm sand >80%, or any Ultisols/Ultic subgroups with 150cm clay <20% and 50cm pH <6.

`"Acid"` is assigned in 4 cases to components that are not Sandy:
1. If the 50cm pH is <5.5 and carbonate depth >150cm and not a Mollisols/Mollic subgroup
2. If the 50cm pH is <6 and carbonate depth >100cm and a Fragi great groups
3. Ultisols and Ultic subgroups
4. Dystr- great groups

`"Calcareous"` is assigned in all remaining cases.

```{r}
#3 Sand
x$Site <- ifelse(grepl("Sandy", x$ecoclassname),
                 yes = "Sandy",
                 no = ifelse((x$T150_sand >= 80 & x$T50_sand >= 70) | 
                                (x$T50_sand >= 80) | 
                                  (x$T150_clay < 20 & x$T50_pH < 6 & grepl('ult', x$taxsubgrp, ignore.case = T)),
                              yes = "Sandy",
                              no = ifelse((x$T50_pH < 5.5 & x$carbdepth > 150 & !grepl('oll', x$taxsubgrp, ignore.case = T)) | 
                                            (x$T50_pH < 6 & x$carbdepth > 100 & grepl('frag', x$taxsubgrp, ignore.case = T)) | 
                                              grepl('ult', x$taxsubgrp, ignore.case = T) | 
                                                grepl('dys', x$taxsubgrp, ignore.case = T), 
                                          yes = "Acid", 
                                          no = "Calcareous")))
```
 
### Water Table depth

The `Water_Table` values are not directly measured properties, but were classes that created in this key:

```{r}
# set all water tables to 250cm (i.e. apprx. out of series control section)
site(x)$Water_Table <- 250

# use hydric rating and drainage class (`hydricrating`, `drainagecl`) to assign values for depth to water
x$Water_Table <- ifelse(x$hydricrating %in% "No" & !is.na(x$hydricrating),
                        yes =  ifelse(x$drainagecl %in% "Moderately well drained" & !is.na(x$drainagecl), 
                                      yes = 100, 
                                      no = ifelse(grepl('Poorly drained', x$drainagecl) & !is.na(x$drainagecl),
                                                  yes = 50, 
                                                  no = x$Water_Table)),
                        no = ifelse(x$hydricrating %in% "Yes" & !is.na(x$hydricrating),
                                    yes = ifelse(x$drainagecl %in% "Very poorly drained" & !is.na(x$drainagecl), 
                                                 yes = 0, 
                                                 no = 25),
                                    no = ifelse(x$drainagecl %in% "Very poorly drained" & !is.na(x$drainagecl), 
                                                yes = 0,
                                                no = ifelse(x$drainagecl %in% "Poorly drained" & !is.na(x$drainagecl), 
                                                            yes = 25,
                                                            no = ifelse(x$drainagecl %in% "Somewhat poorly drained" & !is.na(x$drainagecl), 
                                                                        yes = 50,
                                                                        no = ifelse(x$drainagecl %in% "Moderately well drained" & !is.na(x$drainagecl), 
                                                                                    yes = 100, 
                                                                                    no = 250))))))
```

The final key uses `Site` and `Water_Table` variables calculated in the above two keys to make assignments. 

The key was intended to cover multiple MLRAs, and only output generic groupings. I used a separate table to actually give the ecosite ID to the soil group like a taxonomic synonymy table. 

```{r}
# Subdivide the "Sandy" group into 5 different subgroups
x$Site <- ifelse(x$Site %in% "Sandy",
                 yes = ifelse(x$Water_Table <= 25,
                               yes = "S1 Wet Sandy Depression",
                               no = ifelse(x$Water_Table <= 50,
                                           yes = "S2 Sandy Depression",
                                           no = ifelse(x$Water_Table <= 100,
                                                       yes = "S3 Moist Sandy Plains",
                                                       no = ifelse(x$slope_r >= 15, 
                                                                   yes = "S5 Sandy Slopes", 
                                                                   no = "S4 Sandy Plains")))),
                 no = x$Site)
```

In some MLRAs I lumped groups into the same site concept; rather than rewriting the script, I used the synonymy table to do the job.

```{r}
table(x$Site)
```

### Sand by group

```{r}
plot(density(subset(x, grepl("Sandy", Site))$sandtotal_r, na.rm = TRUE), ylim=c(0,0.1))
lines(density(subset(x, grepl("Calcareous", Site))$sandtotal_r, na.rm = TRUE), col=2, lwd=2, lty=2)
lines(density(subset(x, grepl("Acid", Site))$sandtotal_r, na.rm = TRUE), col=3, lwd=3, lty=2)
legend("topleft", legend = c("Sandy", "Calcareous", "Acid"), lwd=1:3, col=1:3, lty=1:3)
```

### pH by group

```{r}
plot(density(subset(x, grepl("Sandy", Site))$ph1to1h2o_r, na.rm = TRUE), ylim=0:1)
lines(density(subset(x, grepl("Calcareous", Site))$ph1to1h2o_r, na.rm = TRUE), col=2, lwd=2, lty=2)
lines(density(subset(x, grepl("Acid", Site))$ph1to1h2o_r, na.rm = TRUE), col=3, lwd=3, lty=3)
legend("topleft", legend = c("Sandy", "Calcareous", "Acid"), lwd=1:3, col=1:3, lty=1:3)
```
