---
title: Chapter 3 - Exploratory Data Analysis
author: Stephen Roecker and Tom D'Avello
date: "`r Sys.Date()`"
output:
  powerpoint_presentation:
    reference_doc: "../1_introduction/style.pptx"
    fig_caption: yes
    fig_height: 3
    fig_width: 3
editor_options: 
  chunk_output_type: console
bibliography: "../../newbook/book.bib"
biblio-style: apalike
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(warning=FALSE, message=FALSE)

library(aqp)
library(soilDB)
library(ggplot2)
library(dplyr)

```


# Objectives

* Review methods for estimating Low, RV, and High values
* Review different methods for visualing soil data


---


## Range in Characteristics (RIC)

* Why does the L, RV, and H clay represent?
    + What taxonomic unit do they refer too?
* What is the L and H for texture?

```{r, echo=FALSE}

crosby <- fetchSDA(WHERE = "muname LIKE 'Crosby%' AND majcompflag = 'Yes'")
spc <- crosby
c2 <- subsetProfiles(spc, s = paste0("cokey == ", sample(spc, 1)@site$cokey))
c_h <- horizons(c2)
names(c_h) <- gsub("total", "", names(c_h)) 

vars <- c("hzname", "clay_l", "clay_r", "clay_h", "texture")
knitr::kable(c_h[vars], row.names = FALSE)

```


---


##  Descriptive Statistics - Estimates of L, RV & H

```{r, echo=FALSE}

desc <- data.frame(
  Parameter = c("Mean", "Median", "Mode", "Standard Deviation", "Quantiles"),
  NASIS = c(rep("RV", 3), rep("L & H", 2)),
  Description = c("arithmetic average", 
                  "middle value, 50% quantile", 
                  "most frequent value", "variation", 
                  "percent rank of values, such that all values are <= p"
                  ),
  'R function' = c("mean()", "median()", "sort(table(), decreasting = TRUE)[1]", "sd()", "quantile()"),
  check.names = FALSE
  )
knitr::kable(desc)

```
 

# Weighting
## Box example
```{r, weighting}

library(sf)
library(dplyr)
library(terra)
library(sgsR)
library(ggplot2)


# Example ----
# Create a 4 square polygon
n <- 2
bb2 <- st_make_grid(st_bbox(c(xmin = 0, xmax = n, ymin = 0, ymax = n)), n = n)
grd2 <- st_as_sf(bb2)
# grd2$ID <- 1:length(bb2)
grd2$landform = c(rep("backslope", 3), "summit")
# grd3 <- ms_dissolve(grd2, field = "mapunit")
# grd2$comp = c(rep("a", 3), "b")
grd2

n <- 10
bb10 <- st_make_grid(st_bbox(bb2), n = n)
grd10 <- st_as_sf(bb10)
grd10$ID <- 1:length(bb10)
grd10

grd <- terra::intersect(vect(grd2), vect(grd10)) |>
  st_as_sf()

idx <- grd$landform == "backslope"
set.seed(62)
# set.seed(721)
grd <- within(grd, {
  comppct_r <- ifelse(
    idx, 
    rlogis(sum(idx),   location = boot::logit(0.75), scale = 0.5) |> boot::inv.logit(),
    rlogis(sum(! idx), location = boot::logit(0.25), scale = 0.5) |> boot::inv.logit()
    )
  A <- comppct_r
  B <- 1 - comppct_r
  component <- ifelse(comppct_r > 0.5, "A", "B")
  comppct_r <- NULL
})
table(grd$component) |> prop.table()
table(grd$component, grd$landform) |> prop.table(margin = 2) |> round(2)
plot(grd)


# plot box
gg_lf <- ggplot() +
  geom_sf(data = grd, aes(fill = landform)) +
  # ggtitle("map unit") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank()
  ) +
  ggtitle("Landscape Strata")
gg_cn <- ggplot() +
  geom_sf(data = grd, aes(fill = component)) +
  # ggtitle("map unit") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank()
  ) +
  ggtitle("Population/Modeled Class")
gg_a <- ggplot() +
  geom_sf(data = grd, aes(fill = A), ) +
  # ggtitle("map unit") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank()
  ) +
  scale_fill_viridis_c(name = "Component\nA") +
  ggtitle("Population/Modeled %")
gg_b <- ggplot() +
  geom_sf(data = grd, aes(fill = B)) +
  # ggtitle("map unit") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank()
  ) +
  scale_fill_viridis_c(name = "Component\nB") +
  ggtitle("Population/Modeled %")


gridExtra::grid.arrange(gg_cn, gg_lf, gg_b, gg_a, top = "Map Unit A-B")


# stratified sample the grid
set.seed(6912)
test <- mapply(FUN = function(i, x, n) {
  st_coordinates(st_sample(x[[1]][i, ], size = n))
  },
  i = 1:4,
  x = lapply(1:4, function(x) grd2),
  n = c(1, 1, 1, 3),
  SIMPLIFY = FALSE
  )
test <- do.call("rbind", test)
sample <- st_as_sf(as.data.frame(test), coords = 1:2)

ex <- extract(vect(grd), st_coordinates(sample))
sample <- cbind(sample, ex)
sample

sample <- within(sample, {
  # component = c("a", "a", "b", "a", "b", "b")
  size      = c(1, 1, 1, rep(1/3, 3))
  w         = size / 1
})

# simulate clay content and component probabilities
set.seed(2)
sample <- within(sample, {
  # samp_co = c(
  #   # rbinom(3, 1, prob = 0.75),
  #   # rbinom(3, 1, prob = 0.25)
  # )
  samp_clay = c(
    rnorm(3, 15),
    rnorm(3, 30)
  )
})


#   samp_co = ifelse(samp_co == 1, "a", "b")
#   w1 = c(rep(0.75 / 3, 3), rep(0.25 / 3, 3)) * 4
#   # or
#   w2 = c(rep(1/1, 3), rep(1/3, 3))
#   # or 
#   w3 = c(rep(4/1, 3), rep(4/3, 3)) / 4
#   
#   pct  = c(rep(3 / 3, 3), rep(1 / 3, 3)) / 4
#   p = c(rep(1 / 4, 3), rep(3 / 4, 3))  
#   
#   p = 1 / w1
#   w = w1
# })


ggplot() +
  geom_sf(data = grd2, aes(fill = landform)) +
  ggtitle("map unit") +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), axis.title = element_blank()
  ) +
  geom_sf(
    data = sample, #, component = component), 
    aes(shape = component), size = 3
    )
  # geom_sf_text(data = sample, aes(label = samp_co), nudge_y = - 0.1, nudge_x = 0.45) +
  # guides(fill = guide_legend((title = "component")))



# compare tabular summaries
table(pred    = sample$comp, obs       = sample$component) |>
  addmargins()
table(mapunit = sample$comp, component = sample$component) |> 
  addmargins()

vars <- c("component", "comp")
sample[vars] <- lapply(st_drop_geometry(sample[vars]), as.factor)
yardstick::conf_mat(sample, truth = comp, estimate = component, case_weights = w)$table |>
  prop.table(margin = 2) |>
  addmargins()
  

# survey
sample$id <- 1:nrow(sample)
test3 <- sample |>
  srvyr::as_survey_design(
    # probs = 1/w,
    weights = w,
    strata  = landform,
    # ids = id,
    nest = TRUE
    )
srvyr::survey_count(test3, component, vartype = "ci")

(tb <- table(sample$component))
# (wtb <- Hmisc::wtd.table(sample$component, weights = sample$w, type = "table"))
(wtb <- xtabs(w ~ component, data = sample))
(mapunit <- prop.table(wtb) |> round(2))
(accuracy <- mapunit |> as.matrix() |> apply(X = _, 2, max))
c(mapunit, ua = accuracy)
wtb_lf <- xtabs(w ~ landform + component, data = sample) |>
  prop.table() |>
  round(2)
wtb_lf

tb_lf <- xtabs(~ landform + component, data = grd) |> 
  prop.table()
tb_lf |>
  cbind(accuracy = max(tb_lf))
```


### component L-RV-H
```{r}
n <- nrow(sample)
# samp <- st_sample(grd, size = n) |> 
#   st_coordinates() |>
#   terra::extract(x = vect(grd), y = _)
# table(samp$component) |> 
#   prop.table() |> 
#   round(2)
samp_vals6 <- sapply(1:1000, function(i) {
  sample(sample$component, size = n, 
         prob = sample$w,
         replace = TRUE
         ) |> 
    table() |> 
    prop.table()
  }, 
  simplify = FALSE
  ) |> 
  do.call("rbind", args = _)
samp_lrh6 <- apply(samp_vals6, MARGIN = 2, FUN = function(x) quantile(x, p = c(0.1, 0.5, 0.9))) |> 
  t() |> 
  round(2)
samp_lrh6  


n <- 30
samp <- st_sample(grd, size = n) |> 
  st_coordinates() |>
  terra::extract(x = vect(grd), y = _)
table(samp$component) |> 
  prop.table() |> 
  round(2)
samp_vals <- sapply(1:100, function(i) {
  sample(samp$component, n, replace = TRUE) |> 
    table() |> 
    prop.table()
  }, 
  simplify = FALSE
  ) |> 
  do.call("rbind", args = _)
samp_lrh30 <- apply(samp_vals, MARGIN = 2, FUN = function(x) quantile(x, p = c(0.1, 0.5, 0.9))) |> 
  t() |> 
  round(2)
samp_lrh30  


## population values
table(grd$component)
quantile(grd$A, p = c(0.1, 0.5, 0.9))
pop_lrh100 <- grd |> 
  select(A, B) |>
  st_drop_geometry() |>
  apply(X = _, 2, function(x) quantile(x, p = c(0.1, 0.5, 0.9))) |>
  round(2) |>
  t()


# plot component percent ranges
grd_lo <- grd[c("A", "B")] |>
  st_drop_geometry() |>
  stack() 
gg_A <- grd_lo |>
  subset(ind == "A") |>
  ggplot(aes(x = values)) +
  geom_histogram(breaks = seq(0, 1, 0.1)) +
  geom_vline(xintercept = pop_lrh100[1, 1], linetype = "dashed") +
  geom_vline(xintercept = pop_lrh100[1, 3], linetype = "dashed") +
  geom_vline(xintercept = pop_lrh100[1, 2]) +
  # facet_wrap(~ ind, ncol = 1) +
  ggtitle("Component A") +
  xlab("Probability")
gg_B <- grd_lo |>
  subset(ind == "B") |>
  ggplot(aes(x = values)) +
  geom_histogram(breaks = seq(0, 1, 0.1)) +
  geom_vline(xintercept = pop_lrh100[2, 1], linetype = "dashed") +
  geom_vline(xintercept = pop_lrh100[2, 3], linetype = "dashed") +
  geom_vline(xintercept = pop_lrh100[2, 2]) +
  # facet_wrap(~ ind, ncol = 1) +
  ggtitle("Component B") +
  xlab("Probability")
gg_AB_pop <- gridExtra::grid.arrange(gg_A, gg_B, top = "Population Prediction Interval", nrow = 1)

# plot component percent ranges
samp_lo <- samp_vals6 |>
  as.data.frame(col.names = c("A", "B")) |>
  stack() 
gg_As <- samp_lo |>
  subset(ind == "A") |>
  ggplot(aes(x = values)) +
  geom_histogram() +
  geom_vline(xintercept = samp_lrh6[1, 1], linetype = "dashed") +
  geom_vline(xintercept = samp_lrh6[1, 3], linetype = "dashed") +
  geom_vline(xintercept = samp_lrh6[1, 2]) +
  # facet_wrap(~ ind, ncol = 1) +
  ggtitle("Component A") +
  xlim(0, 1) + xlab("Probability")
gg_Bs <- samp_lo |>
  subset(ind == "B") |>
  ggplot(aes(x = values)) +
  geom_histogram() +
  geom_vline(xintercept = samp_lrh6[2, 1], linetype = "dashed") +
  geom_vline(xintercept = samp_lrh6[2, 3], linetype = "dashed") +
  geom_vline(xintercept = samp_lrh6[2, 2]) +
  # facet_wrap(~ ind, ncol = 1) +
  ggtitle("Component B") +
  xlim(0, 1) + xlab("Probability")
gg_AB_samp <- gridExtra::grid.arrange(gg_As, gg_Bs, top = "Hardened Sample Confidence Interval", nrow = 1)
gridExtra::grid.arrange(gg_AB_samp, gg_AB_pop, top = "Map Unit A-B", nrow = 2)


# inverse quantiles, same as hardened frequency proportions 
ecdf_A_t <- ecdf(samp_vals[, 1])
p <- c(0.1, 0.5, 0.9)
rev(1 - ecdf_A_t(p))
table(grd$A > 0.5)



# weighted clay values
weighted.mean(sample$samp_clay, w = sample$w)

sample |>
  st_drop_geometry() |>
  group_by(component) |>
  summarize(wtd_avg = weighted.mean(samp_clay, w = w)) |>
  ungroup()

```

### Accuracy
### Confusion matrix
```{r}
test <- c(30, 70, 10, 90) |>
  matrix(byrow = TRUE, nrow = 2) |>
  as.table()
ua <- apply(test, 1, function(x) max(x)/sum(x)) * 100
pa <- apply(test, 2, function(x) max(x)/sum(x)) * 100
oa <- sum(diag(test)) / sum(test) * 10
cbind(test, ua) |>
  rbind(c(pa, oa)) |>
  round()

```

### RMSE
```{r}
test <- data.frame(index = 1:10, sample = rnorm(5, mean = 10), mean = 10)
test$xmin <- apply(test[-1], 1, min)
test$xmax <- apply(test[-1], 1, max)
test$RV <- as.factor("RV")
test$dif <- test$sample

ggplot(test, aes(x = sample, y = index)) +
  geom_point() +
  geom_vline(xintercept = 10) +
  geom_linerange(aes(xmin = xmin, xmax = xmax), lty = "dashed") +
  ggtitle("Example of the RV value versus sample errors")


```


## Volcano example

```{r}
library(sf)
library(terra)
library(sgsR)
library(geocmeans)


# import volcano DEM, details at http://geomorphometry.org/content/volcano-maungawhau
data("volcano")
volcano_r <- rast(
  volcano[87:1, 61:1],
  crs = crs("+init=epsg:27200"),
  extent = c(
    xmin = 2667405,
    xmax = 2667405 + 61 * 10,
    ymin = 6478705,
    ymax = 6478705 + 87 * 10
  )
)
names(volcano_r) <- "elev"

volcano_r <- rast("C:/workspace2/dem_derivatives2.tif")
nm <- c("z", "hs", "slp", "kp", "kt", "twi", "tpi", "mpi", "rsp", "z_f", "hs_f", "slp_f", "kp_f", "kt_f", "twi_f", "mpi_f")
idx_uf <- nm[c(1, 3:9)]
idx_f  <- nm[c(10, 12:15, 7, 16, 9)]
names(volcano_r) <- nm
v_nm <- names(volcano_r)[c(1, 3:9)]

set.seed(34)
strat_kmeans(volcano_r[[idx_uf]], nStrata = 7) |> 
  freq() |> 
  within({
    p = count / sum(count)
    w = 1/2
    })
set.seed(34)
tab2 <- strat_kmeans(volcano_r[[idx_f]], nStrata = 7) |> 
  freq() |> 
  dplyr::mutate(
    p  = count / sum(count), 
    wp = count / (70 * p),
    wk = count / 10,
    wk / wp,
    1/wk, 
    1/wp,
    1/(1/wk),
    1/(1/wp)
    )

test <- lapply(v_nm, function(n){
  aband <- volcano_r[[n]]
  aband$nrows <- nrow(aband)
  return(aband)
})
names(test) <- v_nm


fcm_r <- CMeans(test, k = 7, m = 1.6, standardize = TRUE)

fcm_r2 <- fcm_r$rasters |>
  rast()
idx <- seq(1, nlyr(fcm_r2), 2)
fcm_r2 <- fcm_r2[[idx]]

nm <- c("Alpha", "Beta", "Gamma", "Delta", "Epsilon", "Zeta", "Eta", "k")
names(fcm_r2) <- nm
levels(fcm_r2$k) <- data.frame(id = 1:7, compname = nm[1:7]) 
summary(fcm_r2[[-8]] |> max())
fcm_r2 |> plot()

names(fcm_r2)[8] <- "k_0p025acres"
k_smooth5 <- fcm_r2[[1:7]] |> bayesEO::bayes_smooth(window_size = 5) |>
  bayesEO::bayes_label()
# fcm_r2$k_smooth5_0p1acres <- sieve(fcm_r2[[9]], threshold = 4,   direction = 4)
fcm_r2$k_smooth5_1p4acres <- sieve(k_smooth5, threshold = 58,  direction = 4)
# fcm_r2$k_smooth5_5p7acres <- sieve(fcm_r2[[9]], threshold = 225, direction = 4)
plot(fcm_r2)
fcm_r <- fcm_r2
names(fcm_r)[8:9] <- c("obs", "map")

sgsR::allocate_existing_equal(fcm_r2[[8]])


# distance
fcm_dist <- fcm_r$Centers |> 
  as.data.frame() |> 
  dist(upper = TRUE) |>
  as.matrix()

comb <- expand.grid(2:7, 1:7) |>
  apply(X = _, 2, as.character) |>
  as.data.frame() |>
  subset(Var1 != Var2)
idx <- apply(comb, 1, function(x) paste(sort(x), collapse = ", ")) |> duplicated()
comb <- comb[!idx, ]
comb$dist <- NA
for (i in 1:nrow(comb)) {
  # add cross-covariance
  comb[i, "dist"] <- fcm_dist[comb[i, 1], comb[i, 2]]
}

```

### sample strata
```{r}
# sample strata
names(fcm_r2)[8:9] <- c("obs", "map")
strata <- fcm_r2[[8]]
names(strata) <- "strata"
test_samp1 <- sample_strat(strata, 10, allocation = "equal", method = "random")
samp <- extract(fcm_r2, st_coordinates(test_samp1))

# sample final map
names(fcm_r2)[8:9] <- c("obs", "map")
strata <- fcm_r2[[9]]
test_samp2 <- spatSample(strata, size = 10, method = "stratified", as.points = TRUE)
samp <- extract(fcm_r2, test_samp2)[9:10]

table(map = samp$map, obs = samp$obs)

population <- as.data.frame(fcm_r2)[8:9]
table(map = population$map, obs = population$obs)
```


### compute weight
```{r}
# weigts from final map
tab <- freq(fcm_r2[[9]]) |>
# # weights from strata
# tab <- freq(fcm_r2[[8]]) |>
  dplyr::mutate(
    layer = NULL,
    # population
    p  = count / sum(count),
    nk_pop   = count * (70/sum(count)),
    nk_equal = 10,
    wk_pop   = count / nk_pop,
    wk_equal = count / nk_equal,
    pk_pop   = 1 / wk_pop,
    pk_equal = 1 / wk_equal,
    # pr = pk_p / pk_e,
    wk_ratio = wk_equal / wk_pop
    # nr = nk_p / nk_e
  )
idx <- c(4:11)
tab[idx] <- lapply(tab[idx], function(x) round(x, 2))
tab

cor(tab$pr, tab$wr)
cor(tab$wk_equal, tab$wk_ratio)
```


### accuracy
```{r}

# weigts from final map
sample <- merge(samp, tab, by.x = "map", by.y = "value")
# # weigts from strata
# sample <- merge(samp, tab, by.x = "obs", by.y = "value")

save(
  sample = sample, 
  population = population,
  rasters = fcm_r,
  file = "C:/workspace2/github/ncss-tech/stats_for_soil_survey/data/exploratory_data/volcano.RData"
  )


# weighted confusion matrix
(cm2 = table(map = sample$map, obs = sample$obs))
cm <- xtabs(wk_equal ~ map + obs, data = sample)
cm_scaled <- round(cm/70, 1)
cm_scaled


# map unit percentages
comppct_r <- prop.table(cm_scaled, margin = 1) |>
  round(2)*100
comppct_r


# confusion matrix metrics
cm_metrics <- aqp::tauW(cm_scaled)
cm_metrics$overall.naive |> round(2)
cm_metrics$user.naive |> round(2)
cm_metrics$prod.naive |> round(2)

cbind(as.data.frame.matrix(cm_scaled), ua = cm_metrics$user.naive) |>
  round(2) |> 
  rbind(pa = c(cm_metrics$prod.naive, cm_metrics$overall.naive) |> round(2))


# population map proportions
table(map = population$map, obs = population$obs) |> 
  prop.table(1) |> 
  round(2) |>
  addmargins()*100


cm_p <- round(prop.table(cm, margin = 1) * 100)

nm <- colnames(cm)
nm_new <- apply(cm_p, 1, function(x) {
  idx <- which(x > 15)
  idx <- idx[order(x[idx], decreasing = TRUE)]
  paste0(nm[idx], collapse = "-")
})
rownames(cm_p) <- nm_new

ua <- round(diag(cm) / apply(cm, 1, sum), 2) * 100
pa <- round(diag(cm) / apply(cm, 2, sum), 2) * 100
oa <- round(sum(diag(cm)) / sum(cm) * 100)

cm2 <- cbind(cm_p, ua)
cm2 <- rbind(cm2, pa = c(pa, oa))

dimnames(cm2) <- list(
  obs  = dimnames(cm2)[[1]],
  pred = dimnames(cm2)[[2]]
)
cm2
mapac::aa_card(x = sample$obs, m = sample$map)
mapac::aa_stratified(
  stratum   = sample$map,
  reference = sample$obs,
  map       = sample$map,
  h   = sort(levels(sample$map)),
  N_h = rep(10, 7)
)


# survey
library(survey)

volcano_svy <- svydesign(ids = ~1, weights = ~wk_equal, stratum = ~pred, data = sample, variables = ~map + obs)
tb <- svytable(~map + obs, design = volcano_svy) |>
  svyciprop()


# srvyr
library(srvyr)

ds <- srvyr::as_survey_design(
  sample,
  weights = wk_equal,
  strata = map
)
ds |>
  group_by(map, obs) |>
  summarize(p = survey_prop(vartype = "ci", level = 0.95, prop_method = "logit")) |>
  filter(map == "Alpha") # first map unit


# DescTools
cm_map_CI <- cm_scaled |> 
  apply(MARGIN = 1, FUN = function(x) {
    DescTools::MultinomCI(x, conf.level = 0.95, method = "wilson") |>
      round(2)
    }, 
    simplify = FALSE
    )
cm_map_CI$Alpha # first map unit


# bootstrap
test <- by(sample, sample$map, function(x){
  sapply(1:1000, function(i){
    sample(x$obs, size = length(x), replace = TRUE, prob = x$wk_e) |>
      table() |>
      prop.table()
  },
  simplify = FALSE
  ) |>
    do.call("rbind", args = _)
})
test <- lapply(test, function(x){
  apply(x, 2, function(y){
    quantile(y, probs = c(0, 0.5, 1)) |> round(2)
  })
})
test$Alpha

samp_vals6 <- sapply(1:1000, function(i) {
  sample(sample$component, size = n, 
         prob = sample$w,
         replace = TRUE
         ) |> 
    table() |> 
    prop.table()
  }, 
  simplify = FALSE
  ) |> 
  do.call("rbind", args = _)

```


---


# Descriptive Statistics - Numeric Data

```{r}
data("loafercreek")
h <- horizons(loafercreek) 
h$texture_class <- factor(h$texture_class)

h %>%
  select(clay, phfield, total_frags_pct, texture_class) %>%
  summary()
```


---


## Descriptive Statistics - Categorical Data

```{r}

table(h$genhz, h$texture_class, useNA = "ifany")

```


---


## Data Inspection - Missing Data

 1. **Exclude** all rows or columns that contain missing values using the function `na.exclude()`, such as `h2 <- na.exclude(h)`. However this can be wasteful because it removes all rows (e.g., horizons), regardless if the row only has 1 missing value. Instead it's sometimes best to create a temporary copy of the variable in question and then remove the missing variables, such as `clay <- na.exclude(h$clay)`.
 2.	**Replace** missing values with another value, such as zero, a global constant, or the mean or median value for that column, such as `h$clay <- ifelse(is.na(h$clay), 0, h$clay) # or h[is.na(h$clay), ] <- 0`.
 3. **Read** the help file for the function you're attempting to use. Many functions have additional arguments for dealing with missing values, such as `na.rm`.


---


## Confidence Interval
```{r}
set.seed(342)
test <- lapply(1L:10L, function(x) data.frame(sample = paste("sample", x), value = rnorm(100, mean = 0))) |>
  do.call("rbind", args = _) |>
  as.data.frame()
test$sample <- factor(test$sample, levels = paste("sample", 1:10))
pop <- data.frame(sample = "population", value = qnorm(seq(0.01, 0.99, 0.01)))
test2 <- test |>
  group_by(sample) |>
  summarize(
    mean = mean(value),
    L1 = DescTools::MeanCI(value)[2],
    H1 = DescTools::MeanCI(value)[3]
    ) |>
  mutate(type = "95% confidence interval")
test3 <- test |>
  group_by(sample) |>
  summarize(
    mean = mean(value),
    L2 = quantile(value, probs = 0.1),
    H2 = quantile(value, probs = 0.9)
    ) |>
  mutate(type = "10th to 90th quantile")
ggplot(test2, aes(x = sample)) + 
  geom_point(aes(y = mean)) +
  geom_linerange(aes(y = mean, ymin = L1, ymax = H1, col = type), size = 3, alpha = 0.5) +
  geom_linerange(data = test3, aes(ymin = L2, ymax = H2, col = type)) +
  geom_hline(yintercept = 0, lty = "longdash", alpha = 0.5) +
  coord_flip() +
  geom_boxplot(data = pop, aes(y = value, x = "population")) + 
  xlab(NULL) + ylab("values") +
  ggtitle("Several confidence intervals estimated from the same population")

```

# Graphical Methods - Descriptions

```{r, echo=FALSE}

figs <- data.frame(
  'Plot Types' = c("Bar", "Histogram", "Density", "Quantile-Quantile", "Box-Whisker", "Scatter & Line"),
   Description = c("a plot where each bar represents the frequency of observations for a 'group'",
                   "a plot where each bar represents the frequency of observations for a 'given range of values'",
                  "an estimation of the frequency distribution based on the sample data",
                  "a plot of the actual data values against a normal distribution",
                  "a visual representation of median, quartiles, symmetry, skewness, and outliers",
                   "a graphical display of one variable plotted on the x axis and another on the y axis"
                  ),
  check.names = FALSE
  )

knitr::kable(figs)

```


---
 

## Graphical Methods - Functions

```{r, echo=FALSE}

figs <- data.frame(
  'Plot Types' = c("Bar", "Histogram", "Density", "Quantile-Quantile", "Box-Whisker", "Scatter & Line"),
  # Description = c("a bar plot where each bar represents the frequency of observations for a given range of values",
  #                "an estimation of the frequency distribution based on the sample data",
  #                "a plot of the actual data values against a normal distribution",
  #                "a visual representation of median, quartiles, symmetry, skewness, and outliers",
  #                 "a graphical display of one variable plotted on the x axis and another on the y axis",
  #                "plots formatted for the representation of circular data"
  #                ),
  'Base R' = c("barplot()", "hist()", "plot(density())", "qqnorm()", "boxplot()", "plot()"),
  'lattice' = c("barchart()", "histogram()", "densityplot()", "qq()", "bwplot()", "xyplot"),
  'ggplot geoms' = c("geom_bar()", "geom_histogram()", "geom_density()", "geom_qq()", "geom_boxplot()", "geom_point()"),
  check.names = FALSE
  )

knitr::kable(figs)

```


---


# Distributions

```{r, echo=FALSE, fig.dim = c(8, 4)}

data("metadata")

# Generalized the horizon designations
n <- c("A",
       "BAt",
       "Bt1",
       "Bt2",
       "Cr",
       "R")
# REGEX rules
p <- c("A",
       "BA|AB",
       "Bt|Bw",
       "Bt3|Bt4|2B|C",
       "Cr",
       "R")

loafercreek$genhz <- generalize.hz(loafercreek$hzname, n, p)

h <- horizons(loafercreek)
h <- within(h, {
  hzdepm = (hzdept + hzdepb) / 2
  texture_class = factor(texture_class, 
                         levels = metadata[metadata$ColumnPhysicalName == "texcl", "ChoiceName"]
                         )
  })

idx <- is.na(h$texture_class)

# bar plot
p_b <- ggplot(h[!idx, ], aes(x = texture_class)) +
  geom_bar() +
  ylab("count (n)") + xlab("texture class") +
  theme(aspect.ratio = 1) +
  ggtitle("Bar Plot")

# histogram
p <- ggplot(data = h, aes(x = clay))

p_h <- p + 
  geom_histogram() + 
  xlab("clay (%)") + ylab("count (n)") +
  theme(aspect.ratio = 1) +
  ggtitle("Histogram")

# density plot
p_d <- p + 
  geom_density(fill = "grey", alpha = 0.5) + 
  xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Density Plot")

gridExtra::grid.arrange(p_b, p_h, p_d, ncol = 3)

```


---


## The 'Normal' Distribution

```{r, echo = FALSE, fig.dim = c(8, 4)}

dnorm_limit_1 <- function(x) {
  y <- dnorm(x)
  y[-1 > x | x > 1] <- NA
  return(y)
  }
dnorm_limit_2 <- function(x) {
  y <- dnorm(x)
  y[-2 > x | x > 2] <- NA
  return(y)
  }

# ggplot() with dummy data
ggplot(data.frame(x=c(-3, 3)), aes(x=x)) + 
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm_limit_1, geom="area", fill="blue", alpha=0.2) +
  stat_function(fun=dnorm_limit_2, geom="area", fill="orange", alpha=0.2) +
  annotate("text", x = 0, y = 0.2, label = "1 sd = 68%") +
  annotate("text", x = -1.5, y = 0.03, label = "2 sd = 95%")

```


---


## Mean & SD vs Median & Quantiles


```{r, echo=FALSE, fig.dim = c(8, 4)}

p   <- c(0.025, 0.5, 0.975)

avg <- mean(h$clay, na.rm = TRUE)
std <- sd(h$clay,   na.rm = TRUE)

clay <- rbind(
  data.frame(
    value = c(avg - 2 * std, avg, avg + 2 * std),
    variable = "mean & 2SD",
    stringsAsFactors = TRUE
    ),
  data.frame(
    value    = quantile(h$clay, p, na.rm = TRUE),
    variable = "median & quantiles",
    stringsAsFactors = TRUE
    )
  )

avg <- mean(h$total_frags_pct, na.rm = TRUE)
std <- sd(h$total_frags_pct,   na.rm = TRUE)

frags <- rbind(
  data.frame(
    variable = "mean & 2SD",
    value = c(avg - 2 * std, avg, avg + 2 * std),
    stringsAsFactors = TRUE
    ),
  data.frame(
    variable = "median & quantiles", 
    value    = quantile(h$total_frags_pct, p, na.rm = TRUE),
    stringsAsFactors = TRUE
    )
  )

p1 <- ggplot(h, aes(x = clay)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(data = clay, aes(xintercept = value, lty = variable)) +
  xlab("%") +
  #xlim(-5, max(h$total_frags_pct)) +
  theme(aspect.ratio = 1) +
  ggtitle("Clay")

p2 <- ggplot(h, aes(x = total_frags_pct)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(data = frags, aes(xintercept = value, lty = variable)) +
  xlab("%") +
  #xlim(-5, max(h$total_frags_pct)) +
  theme(aspect.ratio = 1) +
  ggtitle("Rock Fragments")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```


---


## Distributions

```{r, echo=FALSE, fig.dim = c(8, 4)}

# qq plot
p_qq <- ggplot(h, aes(sample = clay)) + 
  geom_qq() +
  geom_qq_line() +
  xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Q-Q Plot")

# box plot
p_box <- ggplot(h, aes(x = 1, y = clay)) + 
  geom_boxplot() +
  ylab("clay (%)") +
  theme(aspect.ratio = 1) +
  coord_flip() +
  ggtitle("Box Plot")

gridExtra::grid.arrange(p_box, p_qq, ncol = 2)

```


---


## Relationships

```{r, echo=FALSE, fig.dim = c(8, 4)}

# scatter plot
p_s <- ggplot(h, aes(x = clay, y = hzdepm)) +
  geom_point() +
  ylim(100, 0) +
  ylab("depth (cm)") + xlab("clay (%)") +
  theme(aspect.ratio = 1) +
  ggtitle("Scatter Plot (Soil Horizons)")

l_p <- ggplot(h) +
  # geom_smooth(aes(y = clay, x = hzdepm), se = FALSE) +
  geom_line(aes(y = clay, x = hzdepm, group = peiid)) +
  xlim(100, 0) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1) +
  ggtitle("Line Plot (Soil Profiles)")

p_l <- ggplot(h) +
  # geom_line(aes(y = clay, x = hzdept, group = peiid)) +
  geom_step(aes(y = clay, x = hzdept, group = peiid), direction = "vh", alpha = 0.5) +
  # geom_smooth(aes(y = clay, x = (hzdept + hzdepb) / 2), se = FALSE) +
  xlim(100, 0) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1) +
  ggtitle("Line Plot")

# l_s <- ggplot(h, aes(y = clay, x = hzdepm, group = peiid)) +
#   geom_step(aes(y = clay, x = hzdept, group = peiid), direction = "vh") +
#   xlim(100, 0) +
#   xlab("depth (cm)") + ylab("clay (%)") +
#   coord_flip() +
#   theme(aspect.ratio = 1) +
#   ggtitle("Step Line Plot (Soil Profiles)")
# 
# test3 <- aqp::slice(loafercreek, seq(0, 100, 2) ~ clay)
# test3 <- horizons(test3)
# test3$hzdepm <- with(test3, (hzdepb + hzdept) / 2)
# 
# l_sm <- ggplot(test3, aes(y = clay, x = hzdept, group = peiid)) +
#   geom_line(stat = "smooth", method = "loess", col = "black", alpha = 0.5) +
#   xlim(100, 0) +
#   xlab("depth (cm)") + ylab("clay (%)") +
#   coord_flip() +
#   theme(aspect.ratio = 1) +
#   ggtitle("Line Plot (Soil Profiles)")


gridExtra::grid.arrange(p_s, p_l, ncol = 2)

```


---


## 3rd Dimension - Color, Shape, Size, Layers, etc...

```{r, echo=FALSE, fig.dim = c(8, 4)}
h$clay2 <- ifelse(is.na(h$clay), 0, h$clay)

ggplot(h[!idx, ], aes(y = clay, x = hzdepm, col = genhz)) +
  geom_point(size = 2) +
  geom_smooth(se = FALSE, col = "black") +
  xlim(100, 0) + ylim(min(h$clay, na.rm = TRUE), max(h$clay, na.rm = TRUE)) +
  xlab("depth (cm)") + ylab("clay (%)") +
  coord_flip() +
  theme(aspect.ratio = 1)

```


---


## 3rd Dimension - Color

```{r, echo=FALSE, fig.dim = c(8, 4)}

h$genhz2 <- factor(h$genhz, levels = rev(levels(h$genhz)))

p_sb1 <- ggplot(h, aes(x = genhz2, fill = texture_class)) +
  geom_bar(position = "stack") +
  theme(aspect.ratio = 1) +
  ggtitle("Stacked Bars") +
  coord_flip()

p_sb2 <- ggplot(h, aes(x = genhz2, fill = texture_class)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  theme(aspect.ratio = 1) +
  ggtitle("Proportional Stacked Bars") + 
  coord_flip()

gridExtra::grid.arrange(p_sb1, p_sb2, ncol = 2)

```


---


## 3rd Dimension - Groups

```{r, echo=FALSE, fig.dim = c(8, 4)}

idx <- h$genhz %in% c("Cr", "R", "not-used")

p_b <- ggplot(h[!idx, ], aes(x = genhz2, y = clay)) + 
  geom_boxplot() +
  ylab("clay (%)") + xlab("genhz") +
  theme(aspect.ratio = 1) +
  coord_flip()

p_d <- ggplot(h[!idx, ], aes(x = clay, col = genhz)) +
  geom_density(alpha = 0.5, lwd = 1.5) +
  xlab("%") +
  theme(aspect.ratio = 1)

gridExtra::grid.arrange(p_b, p_d, ncol = 2)


```


---


## 3rd Dimension - Facets

```{r, echo = FALSE, fig.dim = c(8, 4)}

df <- reshape2::melt(h, 
                     id.vars = c("peiid", "genhz2", "hzdepm"), 
                     measure.vars = c("clay", "phfield", "total_frags_pct")
                     )

idx <- df$genhz2 %in% c("Cr", "R", "not-used")

ggplot(df[!idx, ], aes(y = value, x = genhz2)) +
  geom_boxplot() +
  xlab("genhz") +
  facet_wrap(~ variable, scales = "free_x") +
  coord_flip()
```


---


## 3rd Dimension - Facets

```{r, echo=FALSE, fig.dim=c(8, 4)}

data("loafercreek", package = "soilDB")

s <- slab(loafercreek, 
          fm = ~ clay + phfield + total_frags_pct, 
          slab.structure = 0:100,
          slab.fun = function(x) quantile(x, c(0.1, 0.5, 0.9), na.rm = TRUE)
          )
names(s) <- gsub("\\.", "", names(s))
names(s) <- gsub("^X", "p", names(s))

s$idx <- s$top %in% seq(0, 100, 10)

cf <- s %>%
  group_by(variable) %>%
  summarize(
    dep = top,
    max = max(p90, na.rm = TRUE) - (diff(range(p90, na.rm = TRUE)) * 0.05),
    cf  = paste0(round(contributing_fraction * 100), "%"),
    idx = idx
    ) %>%
  ungroup() %>%
  filter(idx == TRUE)

ggplot(s, aes(x = top, y = p50)) +
  geom_line() +
  geom_ribbon(aes(ymin = p10, ymax = p90, x = top), alpha = 0.2) +
  geom_text(data = cf, aes(x = dep, y = max, label = cf)) +
  xlim(c(100, 0)) + xlab("depth (cm)") + ylab("") +
  coord_flip() +
  facet_wrap(~ variable, scales = "free_x") +
  ggtitle("Loafercreek")

```


---


## 3rd Dimension - Correlation

```{r, echo=FALSE, fig.dim = c(8, 4)}

h <- transform(h, 
               frags = total_frags_pct,
               depth = hzdepm
               )
vars <- c("depth", "clay", "phfield", "frags")

GGally::ggpairs(h[vars])

```


---


## ggplot2 basics

```{r, eval=TRUE}

h <- horizons(loafercreek)

# set figure axis and aesthetics
p <- ggplot(data = h, aes(x = clay, y = hzdept, color = genhz))

# add layers
p + geom_point()

```


---


## Summary of Components

```{r, echo=FALSE, fig.dim = c(8, 4)}

co <- get_component_from_SDA(WHERE = "compname = 'Miami'")
co$slope_phase <- cut(co$slope_r, c(0, 2, 4, 8, 15, 30, 50))
co2 <- filter(co, majcompflag == "Yes")
h  <- get_chorizon_from_SDA(WHERE = "compname = 'Miami'")
test <- subset(h, cokey %in% co2$cokey)
test <- merge(test, co[c("cokey", "erocl")], by = "cokey", all.x = TRUE)

test2 <- test
test3 <- segment(test2, intervals = 0:100, hzdepcols = c("hzdept_r", "hzdepb_r"))
test3$hzdepm_r <- with(test3, (hzdepb_r + hzdept_r) / 2)

hvars <- c("claytotal_r", "om_r", "caco3_r")
pvars <- c("erocl")
test4 <- reshape(test3[c("cokey", "hzdept_r", hvars, pvars)],
                direction = "long",
                timevar = "variable", times = hvars,
                v.names = "value",    varying = hvars
                )
test4$variable <- factor(test4$variable, levels = c("claytotal_r", "om_r", "caco3_r"))

# ordering of groups is handled by ggplot2
ggplot(test4, aes(x = hzdept_r, y = value)) +
  geom_step(aes(group = cokey), direction = "vh", alpha = 0.1) +
  # geom_line(aes(group = cokey), alpha = 0.1) +
  geom_smooth(aes(col = erocl), se = FALSE, lwd = 2) +
  xlim(100, 0) +
  xlab("depth (cm)") + ylab("Percent (%)") +
  coord_flip() +
  facet_grid(~ variable, scale = "free_x") +
  labs(col = "Erosion Class") +
  ggtitle("Miami Components")

```


---


## Additional Reading

Healy, K., 2018. Data Visualization: a practical introduction. Princeton University Press. [http://socviz.co/](http://socviz.co/)

Helsel, D.R., and R.M. Hirsch, 2002. Statistical Methods in Water Resources Techniques of Water Resources Investigations, Book 4, chapter A3. U.S. Geological Survey. 522 pages. [http://pubs.usgs.gov/twri/twri4a3/](http://pubs.usgs.gov/twri/twri4a3/)

Kabacoff, R.I., 2018. Data Visualization in R. [https://rkabacoff.github.io/datavis/](https://rkabacoff.github.io/datavis/)

Peng, R. D., 2016. Exploratory Data Analysis with R. Leanpub.  [https://bookdown.org/rdpeng/exdata/](https://bookdown.org/rdpeng/exdata/)  

Wilke, C.O., 2019. Fundamentals of Data Visualization. O'Reily Media, Inc. [https://serialmentor.com/dataviz/](https://serialmentor.com/dataviz/)
