---
title: Pre-course Assignment
author: Dylan Beaudette, Tom D'Avello, Katey Yoast 
output: html_document
html_document:
    keep_md: yes
---

## 1) View Paul Finnel's webinar
https://youtu.be/VcdowqknChQ

## 2) Create a folder on your machine to be used as the working directory for this course

## 3) Open RStudio, verify that version 0.99.467 is installed (Help>About RStudio), and set working directory (Session>Set Working Directory)

## 4) Install and load packages by coping and pasting the following code into the Console window. 
```{r, warning=FALSE, message=FALSE, results="hide"}
##Set CRAN mirror
local({
  r <- getOption("repos")
  r["CRAN"] <- "http://cran.mirrors.hoobly.com/"
  options(repos = r)
})

##Install and load packages and dependencies
ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)) 
      install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE)
  }
packages <- c("aqp", "soilDB", "Rcmdr", "raster", "rms", "randomForest", "shape", "shapefiles", "sp", "spatial", "plyr", "lattice", "circular", "markdown", "SSOAP", "rpart", "sharpshootR", "ggplot2", "rgdal", "knitr", "reshape2", "fBasics", "caret", "dplyr", "printr", "cluster", "ape", "RColorBrewer", "latticeExtra", "plotrix")
ipak(packages)

```

## Establish ODBC to NASIS
https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/soilDB/setup_local_nasis.html?root=aqp

## Get Example Data

## Check Functionality

## Report Back

## Additional Support/Optional Readings
Soil Data Aggregation using R: https://www.youtube.com/watch?v=wD9Y0Qpv5Tw 
Stats for Soil Survey Webinar: https://www.youtube.com/watch?v=G5mFt9k37a4
Introduction to Stats in R: http://www.gardenersown.co.uk/Education/Lectures/R/index.htm#inputting_data



```{r}
sessionInfo()
```

