---
title: Pre-course Assignment
author: Dylan Beaudette, Tom D'Avello, Katey Yoast 
output: html_document
html_document:
    keep_md: yes
---

1. View Paul Finnel's [webinar](https://youtu.be/VcdowqknChQ)

2. Create a folder on your machine to be used as the working directory for this course

3. Open RStudio, verify that version 0.99.467 is installed (Help>About RStudio), and set working directory (Session>Set Working Directory)

4. Install Packages
Copy/paste the following code into the R console window. 
```{r, warning=FALSE, message=FALSE, results="hide"}
##Set CRAN mirror
local({
  r <- getOption("repos")
  r["CRAN"] <- "http://cran.mirrors.hoobly.com/"
  options(repos = r)
})

## Install and packages and dependencies
ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg) > 0) 
      install.packages(new.pkg, dependencies = TRUE)
}

## list of packages
packages <- c("aqp", "soilDB", "sharpshootR", "Rcmdr", "raster", "rms", "randomForest", "shape", "shapefiles", "sp", "spatial", "plyr", "lattice", "circular", "markdown", "httr", "rpart", "ggplot2", "rgdal", "knitr", "reshape2", "fBasics", "caret", "dplyr", "printr", "cluster", "ape", "RColorBrewer", "latticeExtra", "plotrix")

## install
ipak(packages)

## install the latest version of packages from the AQP suite:
install.packages("aqp", repos = "http://R-Forge.R-project.org", type = "source")
install.packages("soilDB", repos = "http://R-Forge.R-project.org", type = "source")
install.packages("sharpshootR", repos = "http://R-Forge.R-project.org", type = "source")

## load packages in the list
sapply(packages, library, character.only = TRUE, quietly=TRUE, logical.return=TRUE)
```

5. Establish [ODBC Connection to NASIS](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/soilDB/setup_local_nasis.html?root=aqp)

6. Get Example Data

7. Check Functionality (do we need this?)

8. Report Back with the output from `sessionInfo()`.

9. Additional Support/Optional Readings
  * [Soil Data Aggregation using R](https://www.youtube.com/watch?v=wD9Y0Qpv5Tw)
  * [Stats for Soil Survey Webinar](https://www.youtube.com/watch?v=G5mFt9k37a4)
  * [Introduction to Stats in R](http://www.gardenersown.co.uk/Education/Lectures/R/index.htm#inputting_data)
  * [AQP Website](http://aqp.r-forge.r-project.org/)


```{r}
sessionInfo()
```

